---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# recoverome

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R package checks](https://github.com/xec-cm/recoverome/actions/workflows/check-bioc.yml/badge.svg)](https://github.com/xec-cm/recoverome/actions/workflows/check-bioc.yml)

**A package in development for studying microbiome recovery after a
perturbation.**

`recoverome` aims to make recovery definitions explicit and reproducible:
what reference a sample is compared with, what counts as a return, how long
that return must persist, and what the sampling schedule can support.

The experimental development version provides `setup_recovery()` to register
a named analysis with explicit episodes, events, and sample membership.
`validate_recovery()` diagnoses the registered structure, changes to consumed
metadata, and the relationship between current and original scope. Reference
estimation, deviation, recovery outcomes, plotting, and extraction remain planned.

Read the [documentation](https://xec-cm.github.io/recoverome/) and the
[architecture contract](https://github.com/xec-cm/recoverome/blob/devel/dev/architecture.md)
for the current scope.

## Why recovery needs its own analysis

A microbiome can approach its initial composition at one visit and move away
at the next. A single pre-intervention sample cannot describe an individual's
usual variation. Long gaps between visits also limit the precision of a
recovery time.

The planned package will distinguish observed return from sustained return
under a stated observation rule. It will keep the reference definition,
sampling coverage, and analysis provenance attached to the data. A lack of
statistical significance against baseline will not be treated as proof of
recovery.

The initial design focuses on longitudinal profiles with a known perturbation
and a prespecified reference. Recovery to a personal baseline describes
similarity to that baseline; it does not establish health or functional
restoration.

## Install the development version

The development branch is `devel`. There is no CRAN or Bioconductor release.

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("xec-cm/recoverome", ref = "devel")
```

## Register an analysis

`setup_recovery()` takes a `TreeSummarizedExperiment` (TSE) and returns it with
a registration in its metadata. It preserves assays and sample annotations.
Here, three samples belong to one episode, whose origin is the start of an
exposure interval. All times are numeric days since enrolment. The package
includes two small synthetic cases: `single_episode` and `repeated_episodes`.
The latter adds a second episode and an explicitly excluded sample. Their
matrices and tables are shared by the executable examples and tests.

```{r registration-inputs}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$single_episode

tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
```

The episode and event tables are supplied explicitly. Registration checks
their identities and references, records the time declaration, and stores
normalized `S4Vectors::DataFrame` tables. It does not select reference samples
or calculate recovery.

```{r register-analysis}
tse <- recoverome::setup_recovery(
  tse,
  analysis_id = "antibiotic",
  episodes = example_data$episodes,
  events = example_data$events,
  time_col = example_data$time_col,
  time_unit = example_data$time_unit,
  time_origin = example_data$time_origin
)
registration <- S4Vectors::metadata(tse)$recoverome$analyses$antibiotic
registration$registration$samples
```

Filtering keeps the original registration snapshot. The retained sample below
is `s3`, while the stored sample scope still contains all three original IDs.

```{r historical-scope}
follow_up <- tse[, "s3", drop = FALSE]
colnames(follow_up)
S4Vectors::metadata(follow_up)$recoverome$analyses$antibiotic$scope$sample_ids
report <- recoverome::validate_recovery(follow_up, analysis_id = "antibiotic")
report$summary
```

The registration remains structurally valid, with unchanged retained metadata
and a reduced sample scope: one of the three registered samples remains. The
report does not modify the object or check assay values. Its diagnostics
separate missing historical observations from changed metadata or broken records.

See the [introductory vignette](https://xec-cm.github.io/recoverome/articles/recoverome.html)
for the stored records and filtering example in more detail. This small
dataset illustrates registration only; it does not establish recovery.

## Available and planned workflow

| Function | Status | Responsibility |
|:---------|:-------|:---------------|
| `setup_recovery()` | Available | Register a named analysis, episodes, and events. |
| `add_reference()` | Planned | Record the reference definition and eligible samples. |
| `add_deviation()` | Planned | Attach deviations from the registered reference. |
| `add_recovery()` | Planned | Attach outcomes under an explicit recovery rule. |
| `recovery_results()` | Planned | Extract results at the requested analysis level. |
| `plot_recovery()` | Planned | Display observations and the recovery definition. |
| `validate_recovery()` | Available | Diagnose registration records, metadata changes, and current scope. |

Analysis IDs match `^[a-z][a-z0-9]*$`. Registration reserves the corresponding
`rec_<analysis>_` prefix but creates no sample result columns. A repeated
analysis ID or an existing column under its prefix is an error; use a new
analysis name to register another analysis.

[RFC 001](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/001-registration-validation.md)
defines registration and its validation report. Filtering does not
enrol new samples or reinterpret the historical record. Statistical fitting,
group comparisons, and recovery-time uncertainty methods require separate
design and validation.

## Development and contributions

The next steps are further preservation checks and separately designed
analytical stages. No benchmark
performance or statistical guarantees are claimed for this version.

See [CONTRIBUTING](.github/CONTRIBUTING.md) for local checks and contribution
guidelines. Please use the [issue tracker](https://github.com/xec-cm/recoverome/issues)
to discuss designs or report a reproducible problem.

Maintainer: Francesc Catala-Moll
([ORCID](https://orcid.org/0000-0002-2354-8648)),
<fcatala@irsicaixa.es>.
