---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.path = "man/figures/README-",
  fig.alt = "Saved deviations and confirmation after removing an intermediate supporting visit."
)
```

# recoverome

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R package checks](https://github.com/xec-cm/recoverome/actions/workflows/check-bioc.yml/badge.svg)](https://github.com/xec-cm/recoverome/actions/workflows/check-bioc.yml)

**Describe microbiome return to a personal baseline under an explicit
observation rule.**

`recoverome` keeps the reference, sample deviations, observed episode outcomes
and their provenance in a `TreeSummarizedExperiment` (TSE). Filtering preserves
the original analysis; validation explains which dependencies can still be
checked against the current data.

A return at one visit can be followed by a rebound. Sparse visits cannot
establish continuous recovery, and similarity to a personal baseline does not
establish health or functional restoration. The package makes these limits
visible alongside the results.

## Install the development version

This is experimental software. The development branch is `devel`, requires
R >= 4.6.0, and has no CRAN or Bioconductor release. Version `0.2.0` is the next
MVP target, not a published release.

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("xec-cm/recoverome", ref = "devel")
```

## From input to observed recovery

This workflow uses a bundled synthetic example: two features, one baseline and
six visits from one participant. No external data download is needed. All
parameters illustrate the method; they are not recommended biological cutoffs.
The [full vignette](https://xec-cm.github.io/recoverome/articles/recoverome.html)
explains the inputs, verifies the numerical results and records session versions.

### Register samples, an episode and its event

The exposure runs from day 10 to day 14 since enrolment. The episode uses its
**start**, day 10, as time zero. Sample IDs are TSE column names; sample
annotations explicitly supply subject, episode and day. The constructor below
also gives the two features their names from the bundled matrix.

```{r register-analysis}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$observed_recovery

tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
tse <- recoverome::setup_recovery(
  tse,
  analysis_id = "observed",
  episodes = example_data$episodes,
  events = example_data$events,
  time_col = example_data$time_col,
  time_unit = example_data$time_unit,
  time_origin = example_data$time_origin
)
```

Setup returns the TSE with a named registration and preserves its assays and
annotations. Use a new analysis name for a different definition; setup and add
operations do not overwrite existing stages. See the accepted
[registration contract](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/001-registration-validation.md).

### Choose a personal reference and calculate deviations

Baseline `b1`, collected before the exposure, has counts `(8, 0)` and composition
`(1, 0)`. The reference is explicit, with no automatic window selection.
`add_deviation()` compares each sample's composition with that fixed reference
using Bray--Curtis dissimilarity, as defined in the
[reference and deviation contract](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/002-personal-baseline-deviation.md).

```{r sample-deviations}
tse <- recoverome::add_reference(
  tse,
  analysis_id = "observed",
  reference = "b1",
  assay = "counts",
  preprocessing = "synthetic counts; no upstream transformations"
)
tse <- recoverome::add_deviation(tse, "observed")
sample_results <- recoverome::recovery_results(tse, "observed", level = "sample")
as.data.frame(sample_results)[, c("sample_id", "relative_time", "deviation")]
```

The values are `0`, `0.75`, `0.25`, `0.125`, `0.125`, `0.5`, `0.125`.
One baseline defines a descriptive reference but cannot estimate usual temporal
variation. With several selected baselines, normalized compositions receive
equal sample weight. The baseline diameter is not a recovery threshold.

### Apply an explicit observation rule

```{r observed-recovery}
rule <- list(threshold = 0.25, persistence = 4, max_gap = 3, horizon = 10)
recovered <- recoverome::add_recovery(tse, "observed", rule)
episode_results <- recoverome::recovery_results(recovered, "observed")
as.data.frame(episode_results)[, c(
  "status", "candidate_time", "confirmation_time", "rebound_time", "coverage"
)]
```

After a deviation above `0.25` at day 0, visits at relative days 2, 4 and 6
are within the band. They span four days with consecutive gaps at most three
days: the candidate is day 2 and confirmation is day 6. The rebound at day 8
does not erase the first confirmation. Coverage reaches the declared 10-day
horizon. This describes observed support, not uninterrupted behavior between
visits. See the accepted
[observed recovery rule](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/003-observed-recovery.md).

### Validate and interpret a filtered analysis

Calculate the complete workflow before filtering required inputs. A subsequent
subset keeps the saved outcomes while removing the corresponding sample values.
Here we remove `s3`, which supported confirmation at relative day 4:

```{r analytical-validation}
recoverome::validate_recovery(recovered, "observed")$summary
filtered <- recovered[, colnames(recovered) != "s3"]
recoverome::validate_recovery(filtered, "observed")$summary

historical <- recoverome::recovery_results(filtered, "observed", scope = "historical")
historical[, c("result_state", "confirmation_time", "validation_complete", "dependencies")]
```

The original analysis has complete validation and unchanged dependencies.
After filtering, confirmation remains day 6, but validation is incomplete and
some dependencies are `not_checked`. An `available` result is a saved result,
not a claim that it was calculated on the current subset. Changed retained
inputs would instead be diagnosed as `changed`; validation never recomputes.

```{r historical-samples}
samples <- recoverome::recovery_results(
  filtered, "observed", level = "sample", scope = "historical"
)
as.data.frame(samples)[, c("sample_id", "relative_time", "result_state", "deviation")]
```

The removed sample retains its identity and registered time, with a missing
deviation. Tables are `S4Vectors::DataFrame` objects; their atomic columns
convert to a data.frame or tibble. Full diagnostics, definitions and supporting
IDs are in `S4Vectors::metadata(historical)$recoverome_view`. This context is a
snapshot and need not survive table conversion. Extract again from the TSE to
refresh it. The default `scope = "current"` selects retained original samples,
or episodes with retained included samples. See the
[extraction and plotting contract](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/004-result-extraction-plotting.md).

### Plot observations and saved evidence

```{r plot-recovery, fig.width=9, fig.height=7}
p <- recoverome::plot_recovery(filtered, "observed", scope = "historical")
p + ggplot2::labs(title = "Historical confirmation after filtering")
```

The separate evidence rail preserves the original milestones. Hollow marks and
dashed supporting spans show missing current evidence, including the removed
intermediate visit; they do not express uncertain timing. Window shading marks
the declared rule, not a confidence interval. The returned ordinary ggplot
object can be customized with ggplot2.

## Optional tidy operations

The core workflow requires no tidy adapter. Optional sample filtering and
reordering through `tidySingleCellExperiment` were verified directly on TSE.
Mutation is limited to ordinary atomic annotations with empty `colData()`
metadata and no column descriptors: the tested adapter loses those two kinds
of annotation information on mutation. The
[tidy interoperability guide](https://github.com/xec-cm/recoverome/blob/devel/dev/tidy-interoperability.md)
records tested operations, versions and unsupported cases. The vignette includes
an optional sample-selection recipe.

## Reproducibility and further work

Run the examples with the installed package in a fresh R session. Record the
session alongside your analysis; the rendered vignette includes this output.
It is omitted here to keep the generated README independent of the build host.

```{r session-info, results='hide'}
utils::sessionInfo()
```

All seven functions used above are implemented. Statistical fitting, group
comparisons and recovery-time uncertainty methods require separate design and
validation. No benchmark performance or statistical guarantees are claimed.
Remaining MVP work and later Bioconductor preparation are tracked in the
[development project](https://github.com/users/xec-cm/projects/10).

See [CONTRIBUTING](.github/CONTRIBUTING.md) for local checks and contribution
guidelines. Please use the [issue tracker](https://github.com/xec-cm/recoverome/issues)
to discuss designs or report a reproducible problem.

Maintainer: Francesc Catala-Moll
([ORCID](https://orcid.org/0000-0002-2354-8648)),
<fcatala@irsicaixa.es>.
