---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# recoverome

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R package checks](https://github.com/xec-cm/recoverome/actions/workflows/check-bioc.yml/badge.svg)](https://github.com/xec-cm/recoverome/actions/workflows/check-bioc.yml)

**A package in development for studying microbiome recovery after a
perturbation.**

`recoverome` aims to make recovery definitions explicit and reproducible:
what reference a sample is compared with, what counts as a return, how long
that return must persist, and what the sampling schedule can support.

The experimental development version provides `setup_recovery()` to register
a named analysis with explicit episodes, events, and sample membership.
`validate_recovery()` diagnoses the registered structure, changes to consumed
inputs or stored results, and the relationship between current and original
scope. It checks the stages that are present without changing the TSE.
`add_reference()` attaches a personal reference from explicitly selected baseline
samples. `add_deviation()` measures sample dissimilarity from those fixed
profiles. `add_recovery()` attaches observed episode outcomes under an explicit
rule. Plotting and extraction remain planned.

Read the [documentation](https://xec-cm.github.io/recoverome/) and the
[architecture contract](https://github.com/xec-cm/recoverome/blob/devel/dev/architecture.md)
for the current scope.

## Why recovery needs its own analysis

A microbiome can approach its initial composition at one visit and move away
at the next. A single pre-intervention sample cannot describe an individual's
usual variation. Long gaps between visits also limit the precision of a
recovery time.

The package distinguishes an observed return from confirmation supported by
later visits under a stated observation rule. It keeps the reference definition,
sampling coverage, and analysis provenance attached to the data. A lack of
statistical significance against baseline will not be treated as proof of
recovery.

The initial design focuses on longitudinal profiles with a known perturbation
and a prespecified reference. Recovery to a personal baseline describes
similarity to that baseline; it does not establish health or functional
restoration.

## Install the development version

The development branch is `devel`. There is no CRAN or Bioconductor release.

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("xec-cm/recoverome", ref = "devel")
```

## Register an analysis

`setup_recovery()` takes a `TreeSummarizedExperiment` (TSE) and returns it with
a registration in its metadata. It preserves assays and sample annotations.
Here, three samples belong to one episode, whose origin is the start of an
exposure interval. All times are numeric days since enrolment. The package
includes small synthetic cases: `single_episode`, `repeated_episodes`, and
`observed_recovery`.
`repeated_episodes` adds a second episode and an explicitly excluded sample;
`observed_recovery` illustrates confirmation followed by a rebound. Their
matrices and tables are shared by the executable examples and tests.

```{r registration-inputs}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$single_episode

tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
```

The episode and event tables are supplied explicitly. Registration checks
their identities and references, records the time declaration, and stores
normalized `S4Vectors::DataFrame` tables. It does not select reference samples
or calculate recovery.

```{r register-analysis}
tse <- recoverome::setup_recovery(
  tse,
  analysis_id = "antibiotic",
  episodes = example_data$episodes,
  events = example_data$events,
  time_col = example_data$time_col,
  time_unit = example_data$time_unit,
  time_origin = example_data$time_origin
)
registration <- S4Vectors::metadata(tse)$recoverome$analyses$antibiotic
registration$registration$samples
```

Filtering keeps the original registration snapshot. The retained sample below
is `s3`, while the stored sample scope still contains all three original IDs.

```{r historical-scope}
follow_up <- tse[, "s3", drop = FALSE]
colnames(follow_up)
S4Vectors::metadata(follow_up)$recoverome$analyses$antibiotic$scope$sample_ids
report <- recoverome::validate_recovery(follow_up, analysis_id = "antibiotic")
report$summary
```

The registration remains structurally valid, with unchanged retained metadata
and a reduced sample scope: one of the three registered samples remains. The
report does not modify the object or check assay values. Its diagnostics
separate missing historical observations from changed metadata or broken records.

See the [introductory vignette](https://xec-cm.github.io/recoverome/articles/recoverome.html)
for the stored records and filtering example in more detail. This small
dataset illustrates the data contracts; it does not establish recovery.

## Attach a personal reference

Select baseline sample IDs explicitly and name the assay. Each selected sample
must precede the start of its episode's origin event, even if the episode uses
that event's end as its time origin. The function closes each selected sample
to proportions over the chosen features, then averages samples equally within
each episode. It performs no implicit filtering, pseudocount addition or
preprocessing beyond that declared closure.

```{r personal-reference}
referenced <- recoverome::add_reference(
  tse,
  analysis_id = "antibiotic",
  reference = "s1",
  assay = "counts",
  preprocessing = "synthetic counts; no upstream transformations"
)
personal <- S4Vectors::metadata(referenced)$recoverome$analyses$antibiotic$reference
personal$profiles
personal$episodes
```

Here the reference is `(0.8, 0.2)`, with `single_sample` support and an undefined
baseline diameter (`NA`), since one observation cannot describe temporal
variation. A reference is descriptive: it is not a healthy-state estimate or
an automatic recovery threshold. Selection and input fingerprints are retained
with the profile; assays and unrelated TSE content remain unchanged.

`validate_recovery()` can check this reference against its saved record and
current baseline inputs. Limited support is reported as information; a single
baseline is not itself a broken analysis.

## Calculate sample deviations

`add_deviation()` uses the assay and feature set recorded by `add_reference()`.
It checks that the selected baselines and their inputs are still available and
unchanged, then compares each included sample composition with its fixed episode
profile using Bray--Curtis dissimilarity.

```{r sample-deviations}
analysed <- recoverome::add_deviation(referenced, analysis_id = "antibiotic")
SummarizedExperiment::colData(analysed)[, c(
  "rec_antibiotic_deviation", "rec_antibiotic_deviation_status"
), drop = FALSE]
```

The deviations are `0`, `0.05`, and `0.4` for `s1`, `s2`, and `s3`. The baseline
sample has zero self-deviation; this is descriptive similarity over the selected
features, not a recovery decision. Included samples without a baseline receive
`NA` and `missing_baseline`; excluded samples receive `NA` and `excluded`.
Neither case is assigned a fabricated distance.

Calculate before removing baseline inputs. Once deviations exist, filtering
keeps the retained sample values and the original analysis history. Reordering
matches identities; it never refits the reference. Repeated additions error,
including identical calls. Use a new analysis name for a different definition.

## Check current dependencies without rewriting history

```{r analytical-validation}
recoverome::validate_recovery(analysed, analysis_id = "antibiotic")$summary

# A saved deviation keeps its value even when current measurements change.
changed <- analysed
SummarizedExperiment::assay(changed, "counts")[, "s3"] <- c(50L, 50L)
changed_report <- recoverome::validate_recovery(changed, analysis_id = "antibiotic")
changed_report$summary
```

The original analysis has unchanged dependencies. In `changed`, the stored
`s3` deviation is still `0.4`, but the report identifies its modified input.
The comparison is complete even though a change was found. By contrast,
removing a required baseline or selected feature leaves its comparison
unavailable and validation incomplete. A subset retains its historical meaning;
validation does not silently recalculate it.

Read `structural_valid`, `validation_complete` and `dependencies` together.
`deviation_status = "computed"` records a past calculation, not present validity.
See the [vignette](https://xec-cm.github.io/recoverome/articles/recoverome.html)
for filtering and output edits, and the
[diagnostic guide](https://github.com/xec-cm/recoverome/blob/devel/dev/analytical-validation.md)
for supported codes and detection boundaries.

## Attach observed recovery outcomes

This seven-sample example uses one explicit baseline and six follow-up visits.
The rule below is illustrative, not a recommended biological threshold; all
four parameters must be supplied and justified for an analysis.

```{r observed-recovery}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$observed_recovery
tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
tse <- recoverome::setup_recovery(
  tse,
  analysis_id = "observed",
  episodes = example_data$episodes,
  events = example_data$events,
  time_col = example_data$time_col,
  time_unit = example_data$time_unit,
  time_origin = example_data$time_origin
)
tse <- recoverome::add_reference(tse, "observed", reference = "b1", assay = "counts")
tse <- recoverome::add_deviation(tse, "observed")
rule <- list(threshold = 0.25, persistence = 4, max_gap = 3, horizon = 10)
recovered <- recoverome::add_recovery(tse, "observed", rule)
outcome <- S4Vectors::metadata(recovered)$recoverome$analyses$observed$recovery
outcome$episodes[, c("status", "candidate_time", "confirmation_time", "rebound_time", "coverage")]
```

The candidate return is observed at relative day 2 and confirmed at day 6.
A rebound at day 8 preserves that first confirmation. Coverage reaches the
10-day horizon; it does not establish uninterrupted recovery between visits.
Outcomes belong to episodes and are stored in metadata, with supporting sample
IDs in `outcome$evidence`. No recovery columns are added to `colData()`.

`add_recovery()` requires the complete realized deviation scope and unchanged
parent dependencies. Calculate before filtering. Later subsets retain the
original outcome; `validate_recovery()` reports missing inputs and any
independently detectable changes without reclassifying it. See the
[vignette](https://xec-cm.github.io/recoverome/articles/recoverome.html) and the
[observed recovery guide](https://github.com/xec-cm/recoverome/blob/devel/dev/observed-recovery.md).

## Available and planned workflow

| Function | Status | Responsibility |
|:---------|:-------|:---------------|
| `setup_recovery()` | Available | Register a named analysis, episodes, and events. |
| `add_reference()` | Available | Attach personal reference profiles, support and input provenance. |
| `add_deviation()` | Available | Attach sample deviations and provenance from the fixed reference. |
| `add_recovery()` | Available | Attach observed episode outcomes under an explicit recovery rule. |
| `recovery_results()` | Planned | Extract results at the requested analysis level. |
| `plot_recovery()` | Planned | Display observations and the recovery definition. |
| `validate_recovery()` | Available | Diagnose analytical dependencies and historical scope. |

Analysis IDs match `^[a-z][a-z0-9]*$`. Registration reserves the corresponding
`rec_<analysis>_` prefix but creates no sample result columns. A repeated
analysis ID or an existing column under its prefix is an error; use a new
analysis name to register another analysis.

[RFC 001](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/001-registration-validation.md)
defines registration and its validation report. Filtering does not
enrol new samples or reinterpret the historical record. Statistical fitting,
group comparisons, and recovery-time uncertainty methods require separate
design and validation.

## Development and contributions

The next steps are recovery integration checks, extraction and plotting under
the accepted contracts. No benchmark performance or statistical
guarantees are claimed for this version.

See [CONTRIBUTING](.github/CONTRIBUTING.md) for local checks and contribution
guidelines. Please use the [issue tracker](https://github.com/xec-cm/recoverome/issues)
to discuss designs or report a reproducible problem.

Maintainer: Francesc Catala-Moll
([ORCID](https://orcid.org/0000-0002-2354-8648)),
<fcatala@irsicaixa.es>.
