---
title: "Diagnose baseline support, sampling and saved outcomes"
author: "Francesc Catala-Moll"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Diagnose baseline support, sampling and saved outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5)
library(recoverome)
```

## Three complementary questions

The diagnostic plots inspect recorded support and observations. They do not
refit a reference, infer a continuous trajectory, rerun a recovery rule or
establish clinical recovery. Use `plot_recovery()` for a detailed deviation plot;
these three views answer narrower questions:

- `plot_reference()`: which baseline samples support the fixed personal profile?
- `plot_sampling()`: when were samples collected, and where are original gaps?
- `plot_recovery_overview()`: which observed milestones and follow-up were saved
  for each subject and episode?

All examples are offline and use the bundled synthetic data. Their illustrative
rule values are explicit choices, not defaults or recommended clinical cutoffs.

```{r register}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$observed_recovery
raw <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
tse <- setup_recovery(
  raw,
  analysis_id = "observed",
  episodes = example_data$episodes,
  events = example_data$events,
  time_col = example_data$time_col,
  time_unit = example_data$time_unit,
  time_origin = example_data$time_origin
)
```

## Recorded reference support

Attach the explicitly selected baseline and inspect it before any deviation
stage exists. Here the single sample is identical to its fixed profile, so its
distance is zero. That supplies no estimate of pairwise baseline variation and
makes no stability claim.

```{r reference, fig.alt="One baseline sample with zero verified distance and no estimable pairwise variation."}
tse <- add_reference(tse, "observed", "b1", assay = "counts")
plot_reference(tse, "observed")
```

The facet summary preserves the recorded sample count, distinct-time count,
support state and diameter. Multiple samples at one time do not supply independent
temporal replication. Individual sample IDs remain visible even when their time
and distance coincide.

Before deviations exist, a distance is derived only from baseline assay values
verified against their stored input fingerprints, using the saved profile. The
legend labels these as current verified baseline distances. Once deviations have
been stored, the plot uses those retained values. It never substitutes a new
measurement after a source edit. Missing or unverifiable values use identity/time
markers below the distance panel, rather than points at zero. The fixed assay,
feature scope and mean-composition definition remain explicit.

## Registered sampling calendar

`plot_sampling()` can be called immediately after registration. Later, a stored
recovery rule adds its horizon and maximum-gap context. The calendar continues to
use the original times even if current annotations are edited.

```{r calculate}
tse <- add_deviation(tse, "observed")
tse <- add_recovery(tse, "observed", list(
  threshold = 0.25, persistence = 4, max_gap = 3, horizon = 10
))
filtered <- tse[, colnames(tse) != "s3"]
```

```{r sampling, fig.alt="Original visits and two-day gaps remain after removing the time-four sample."}
plot_sampling(filtered, "observed", scope = "historical")
```

Hollow markers show removed samples at their original coordinates. The two-day
gaps around the removed time-four sample remain original two-day gaps: filtering
does not manufacture a four-day original gap. Distinct samples collected at one
time stack vertically without moving the underlying time coordinate.

Event intervals and point events are recorded context, not causal attribution.
The solid vertical line marks the chosen event boundary at zero; a dashed line
marks the stored horizon, when present. All consecutive registered-time gaps are
shown, including those outside the recovery window. Highlighting a gap against a
stored cutoff does not re-evaluate an outcome or prove an outside state.

## Saved episode outcomes

The overview distinguishes perturbation, first return, winning candidate,
confirmation, rebound and last observed follow-up. First return and the winning
candidate need not agree. Confirmation can coexist with a later rebound.

```{r overview, fig.alt="Saved confirmation is incomplete after an intermediate supporting sample was removed."}
plot_recovery_overview(filtered, "observed", scope = "historical")
```

Small vertical offsets separate coincident milestones within an episode row;
their times are unchanged. Status and coverage are separate recorded categories.
A confirmed return with incomplete follow-up remains different from complete
observation, and a non-evaluable outcome retains its original reason.

Evidence availability refers to retained IDs. A confirmation marker requires
its entire saved run, including intermediate visits, so removing only time four
makes confirmation evidence incomplete even though the endpoint at time six is
retained. Fully removed evidence is historical. These labels describe current
availability, not uncertainty in the saved milestone's time. Source changes and
incomplete validation are reported separately in the caption.

Subjects and episodes remain explicit when one participant has repeated exposures:

```{r repeated, fig.height=6, fig.alt="Two episodes from the same participant retain separate status and coverage rows."}
example_data <- recovery_examples$repeated_episodes
repeated <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
repeated <- setup_recovery(
  repeated, "repeat", example_data$episodes, example_data$events,
  time_col = example_data$time_col, time_unit = example_data$time_unit,
  time_origin = example_data$time_origin
)
repeated <- add_reference(repeated, "repeat", c("s1", "s4"), assay = "counts")
repeated <- add_deviation(repeated, "repeat")
repeated <- add_recovery(repeated, "repeat", list(
  threshold = 0.25, persistence = 4, max_gap = 3, horizon = 10
))
plot_recovery_overview(repeated, "repeat")
```

A shared relative-time axis does not establish biological comparability across
personal reference definitions. No survival model, confidence interval or
continuous recovery trajectory is estimated by these views.

## Scope and customization

Current scope selects episodes with any retained originally included sample.
Historical scope includes all registered episodes. Within selected episodes,
original observations remain visible with their availability marked. Added
samples are not enrolled and generate the usual structured scope warning.

Empty selections return an informative plot with scope and validation context.
A reference that has not been computed is explicitly labeled; an overview before
recovery retains episodes as not computed. This differs intentionally from
`plot_recovery()`, which requires an eligible episode and deviation stage.

Every function returns an ordinary unprinted ggplot and preserves the input TSE,
its other analyses, provenance and fingerprints. Customize the returned object:

```{r customize, eval=TRUE}
p <- plot_sampling(tse, "observed")
custom <- p + ggplot2::labs(title = "Observation schedule") + ggplot2::theme_bw()
inherits(custom, "ggplot")
```

Printing draws the plot. Customization does not refresh validation; call the
function again against a TSE for a fresh view. See
[history and validation](history-and-validation.html) for the distinction between
saved results and a newly computed analysis after filtering.

## Session information

```{r session}
sessionInfo()
```
