---
title: "Filtering, historical results and validation"
author: "Francesc Catala-Moll"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Filtering, historical results and validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nostos)
```

## Introduction

Calculate the complete synthetic analysis from [Get started](nostos.html).
Its reference is `(1, 0)`; follow-up deviations are `.75, .25, .125, .125, .5,
.125` at relative days 0, 2, 4, 6, 8 and 10. Under the declared rule, candidate,
confirmation and rebound are at days 2, 6 and 8.

```{r original}
data("recovery_examples", package = "nostos")
example_data <- recovery_examples$observed_recovery
raw <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
tse <- setup_recovery(
  raw,
  analysis_id = "observed",
  episodes = example_data$episodes,
  events = example_data$events,
  time_col = "day",
  time_unit = "days",
  time_origin = "days since enrolment"
)
tse <- add_reference(tse, "observed", reference = "b1", assay = "counts",
                     preprocessing = "synthetic counts; no upstream transformations")
tse <- add_deviation(tse, "observed")
rule <- list(threshold = 0.25, persistence = 4, max_gap = 3, horizon = 10)
tse <- add_recovery(tse, "observed", rule)

```

## Remove an intermediate supporting visit

```{r historical-subset}
filtered <- tse[, colnames(tse) != "s3"]
historical <- recovery_results(filtered, "observed", scope = "historical")
as.data.frame(historical)[, c(
  "status", "confirmation_time", "rebound_time", "validation_complete", "dependencies"
)]
report <- validate_recovery(filtered, "observed")
report$summary
stopifnot(historical$confirmation_time == 6, historical$rebound_time == 8,
          !historical$validation_complete, historical$dependencies == "not_checked")
```

Filtering **after calculation** preserves the saved reference, evidence and episode
outcomes. It does not mean that the remaining visits alone could confirm return.
Missing inputs make some comparisons unavailable; the historical outcome remains
readable. For a new question, filter raw inputs **before registering a new name**:

```{r new-subset-analysis}
short <- raw[, colnames(raw) != "s3"]
short <- setup_recovery(
  short, "short", example_data$episodes, example_data$events,
  time_col = "day", time_unit = "days", time_origin = "days since enrolment"
)
short <- add_reference(short, "short", "b1", assay = "counts")
short <- add_deviation(short, "short")
short <- add_recovery(short, "short", rule)
new_result <- recovery_results(short, "short")
new_result[, c("status", "candidate_time", "confirmation_time")]
stopifnot(new_result$status == "unconfirmed_return", is.na(new_result$confirmation_time))
```

Without day 4, the gap from day 2 to day 6 is four days, exceeding the allowed
three. The new analysis has an unconfirmed return. Adding recovery to an old
deviation stage after dropping inputs is unsupported: start a new analysis.
New and historical names can coexist on the same TSE without rewriting each other.

```{r historical-plot, fig.width=8, fig.height=6, fig.alt="Historical confirmation persists after day 4 is removed; the evidence rail marks missing supporting evidence."}
plot_recovery(filtered, "observed", scope = "historical")
```

Hollow marks and dashed support on the evidence rail indicate missing current
evidence. They do not represent timing uncertainty. Saved outcomes and current
measurements occupy separate visual layers.

## Remove a baseline or an entire episode

```{r missing-baseline}
without_baseline <- tse[, colnames(tse) != "b1"]
validate_recovery(without_baseline, "observed")$summary
stopifnot(recovery_results(without_baseline, "observed")$confirmation_time == 6)
```

Removing `b1` does not refit the reference or erase confirmation. Its input can
no longer be compared. A newly registered analysis with no eligible baseline
would instead have `missing_baseline` support and a `not_evaluable` outcome.

This one-episode example loses the whole episode when all its samples disappear:

```{r missing-episode}
empty <- tse[, FALSE]
current <- recovery_results(empty, "observed")
old <- recovery_results(empty, "observed", scope = "historical")
stopifnot(nrow(current) == 0L, nrow(old) == 1L,
          !old$current_present, old$confirmation_time == 6)
old[, c("episode_id", "current_present", "result_state", "confirmation_time")]
```

The current table is typed and empty. Historical scope retains the episode and
its saved evidence. In a multi-episode analysis, the same rule applies to each
episode: a current episode needs at least one retained included sample.

## Change a retained input

```{r changed-input}
changed <- tse
SummarizedExperiment::colData(changed)$day[colnames(changed) == "s4"] <- 30
changed_report <- validate_recovery(changed, "observed")
changed_report$summary
stopifnot(changed_report$summary$dependencies == "changed")
saved <- recovery_results(changed, "observed", level = "sample")
stopifnot(saved$relative_time[saved$sample_id == "s4"] == 6)
```

A changed current day does not rewrite the registered day. Extraction returns
the original coordinates with validation flags. Changes to consumed assay values
are also checked once analytical stages exist. Even positive rescaling of raw
abundances changes their source fingerprint despite unchanged compositions.
Registration alone does not consume assays; unrelated metadata and unselected
measurements are outside that stage's dependency checks.

## Read validation fields together

| Field | Interpretation |
|:---|:---|
| `structural_valid` | FALSE for a demonstrated malformed record, container or authoritative output; NA if unsupported content prevents a conclusion and no failure is known; TRUE otherwise. |
| `validation_complete` | Whether all required comparisons could run. A fully checked source can still be changed. |
| `dependencies` | `changed` when a relevant difference is known; otherwise `not_checked` if a comparison is unavailable; `unchanged` only when the required comparisons agree. |

Removing a baseline and editing a retained owned output can yield both missing
input and output inconsistency findings. Unavailable checks do not suppress
independent failures. Read `report$diagnostics` for components and IDs; message
wording and row order are not contracts. `deviation_status = "computed"` records
a historical computation, not current agreement or biological validity.

Sample/feature scope is separate: reorder is `same`; removal is `subset` or
`empty`; additions give `expanded` or `mixed`. Renaming an ID means removal plus
addition; equal values or tree links do not establish identity. New identities
are never enrolled automatically. Reference-only validation can remain complete
after removing an unused feature or a non-baseline sample; later stages consume
more inputs and therefore have more historical dependencies.

Validation never repairs metadata, refits, reclassifies or changes timestamps.
It checks stored invariants and fingerprints without estimating the reference
again. Unknown schemas/stages/fingerprint formats remain incompletely checked,
not falsely certified as unchanged. Fingerprints detect ordinary edits, not an
adversarial rewrite of all records and hashes. Manual metadata editing is unsupported.

## Extract tables without losing context

```{r sample-availability}
samples <- recovery_results(filtered, "observed", level = "sample", scope = "historical")
as.data.frame(samples)[, c("sample_id", "relative_time", "result_state", "deviation")]
stopifnot(samples$result_state[samples$sample_id == "s3"] == "removed",
          is.na(samples$deviation[samples$sample_id == "s3"]))
```

Rows follow original registration order. `available` means a readable saved result,
including valid missing-baseline or excluded statuses; `removed` sample deviations
and statuses are missing because hashes cannot reconstruct them. `not_computed`
means the relevant stage or realized sample result does not exist. Saved episode
outcomes remain available after their samples disappear. Excluded sample identities
are retained with missing subject, episode and time fields.

Tables are `S4Vectors::DataFrame` objects with atomic columns. `as.data.frame()`
and optional tibble conversion give ordinary tables, but do not promise to retain
context. Validation flags describe the **whole selected analysis**, not individual
rows. Editing a returned table does not update or revalidate the TSE.

```{r context}
context <- S4Vectors::metadata(samples)$recoverome_view
names(context)
context$validation$summary
```

Context contains the selection, full validation report, registered time/event
information, original and realized scopes, reference/deviation/recovery definitions,
provenance and fingerprints, and full saved evidence IDs. Absent stages have NULL
entries; unsupported optional stages are listed as uninterpreted. Extract again
from the TSE for fresh context. Definitions and evidence can remain historical
even when a current table has fewer rows.

Readable results survive unavailable or changed inputs. Ambiguous identities,
malformed required records, broken parent/self fingerprints or corrupted retained
outputs prevent coherent extraction and error. Some unsupported optional stages
may remain opaque when the selected level does not need them. See
`?recovery_results` and the accepted
[extraction contract](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/004-result-extraction-plotting.md)
for the exact schema and failure policy.

## Container and practical limits

Recoverome uses public TSE accessors and preserves unrelated assays, annotations,
trees and links supplied to its own calls. Native subsetting is performed by TSE;
empty axes or unrepresented trees may lose trees/links through the container's own
behavior. Tested bracket selection does not guarantee arbitrary aggregation,
renaming, tree editing or adapter transformations.

Analytical validation and extraction may materialize relevant assay blocks.
They make no general memory/I/O guarantee for arbitrary matrix backends. The
[optional tidy assessment](https://github.com/xec-cm/recoverome/blob/devel/dev/tidy-interoperability.md)
records tested versions and operations: ordinary annotation mutation in the tested
adapter loses colData metadata and column descriptors, so it is unsupported when
those must be preserved. No adapter is needed for this vignette.

## Diagnostic reference

| Code | Severity | Meaning |
|:-----|:---------|:--------|
| `REFERENCE_RECORD_INVALID`, `DEVIATION_RECORD_INVALID`, `RECOVERY_RECORD_INVALID` | error | A known record has malformed fields or inconsistent identities. |
| `SCHEMA_UNSUPPORTED` | error | A record version is unknown; its inner layout is not interpreted. |
| `STAGE_UNSUPPORTED` | error | A future stage, or unknown owned content, is outside the implemented checks. |
| `FINGERPRINT_FORMAT_UNSUPPORTED` | error | The format or current encoding compatibility cannot support fingerprint comparisons. This alone is not a detected data change. |
| `FINGERPRINT_CHANGED` | error | A stored record or its parent fingerprint does not agree. |
| `ANALYTICAL_INPUT_CHANGED` | error | A consumed current source differs from its recorded dependency. |
| `ASSAY_MISSING`, `ASSAY_AMBIGUOUS` | error | The recorded assay cannot be selected uniquely. |
| `ASSAY_VALUES_INVALID` | error | A consumed source no longer has accepted numeric values. |
| `ASSAY_READ_FAILED` | error | The required assay block cannot be read or converted; its comparison remains unavailable. |
| `RESULT_INCONSISTENT` | error | A retained owned output is absent, malformed, or inconsistent with its recorded hash. |
| `REFERENCE_INPUT_MISSING` | info | Baseline samples or selected features needed for current reference comparison were removed. |
| `DEVIATION_SAMPLE_MISSING` | info | A sample in the recorded deviation scope is no longer available for comparison. |
| `RECOVERY_INPUT_MISSING` | info | A recorded recovery input is absent; outcomes retain their historical scope. |
| `BASELINE_SUPPORT_LIMITED` | info | Missing baseline, one sample or one sampling time limits interpretation; this is not itself a structural error. |


See the [developer evidence map](https://github.com/xec-cm/recoverome/blob/devel/dev/README.md)
for focused integration tests, and the [real-data example](real-data.html) for
scientific limits. Technical consistency does not establish a valid biological
threshold, uncertainty interval or clinical interpretation.

## Session information

```{r session-info}
utils::sessionInfo()
```
