---
title: "Prepare longitudinal inputs and an explicit recovery rule"
author: "Francesc Catala-Moll"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Prepare longitudinal inputs and an explicit recovery rule}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(recoverome)
```

## Introduction

A TSE has features in rows and samples in columns. Use unique, stable row and
column names; sample metadata must align with those column names. Give the assay
an explicit name and describe its preprocessing. Counts or nonnegative normalized
abundances are accepted; negative log-transformed values are not. Selected samples
need finite, nonnegative values and a positive total over selected features.
Recoverome closes those values to compositions; it does not normalize sequencing
bias, infer absolute abundance, add pseudocounts or select taxa automatically.

```{r repeated-input}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$repeated_episodes
tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
SummarizedExperiment::colData(tse)
example_data$episodes
example_data$events
```

This example has **two episodes in the same participant**. Each included sample
belongs to exactly one episode, and each episode belongs to one subject. Sample
`s6` has no episode and remains excluded. Excluded samples stay in the TSE but do
not supply a reference or outcome. Every supplied episode needs at least one
included sample at registration. Explicit subject and episode identities prevent
accidentally combining repeated exposures or borrowing another person's baseline.

## Keep event boundaries and time units explicit

Events have an `event_id`, `episode_id`, `start_time` and `end_time`. Equal start
and end represents a point event; an interval represents a duration. Each episode
selects an origin event and either its `start` or `end` boundary as time zero.
In this fixture episode `e1` uses day 10 (start), whereas `e2` uses day 42 (end).
Baseline eligibility always uses **exposure start**, even for an end origin.

```{r register}
tse <- setup_recovery(
  tse, "courses", example_data$episodes, example_data$events,
  time_col = "time", time_unit = "days", time_origin = example_data$time_origin
)
registered <- recovery_results(tse, "courses", level = "sample")
as.data.frame(registered)[, c("sample_id", "episode_id", "included", "relative_time")]
stopifnot(identical(registered$relative_time, c(-7, 0, 7, -8, 1, NA_real_)))
```

Times are finite numeric coordinates, not Date/POSIXct columns. Convert dates
explicitly before registration, using the same origin for samples and events.
The illustrative conversion below does not change the fixture:

```{r calendar-conversion}
enrolment <- as.Date("2025-01-01")
sample_dates <- as.Date(c("2025-01-04", "2025-01-11", "2025-01-18"))
as.numeric(difftime(sample_dates, enrolment, units = "days"))
```

If participants have different enrolment dates, convert within participant.
Document timezone and duration conventions when converting timestamps. A time
unit string records your choice; the package does not convert it or infer dates.
Events from another episode of the **same subject** can block recovery when they
overlap the evaluation window, even after an apparent early confirmation. This
avoids silently attributing observations to one of overlapping exposures.

## Select a defensible personal baseline

```{r reference}
tse <- add_reference(tse, "courses", reference = c("s1", "s4"), assay = "counts",
                     preprocessing = "bundled synthetic counts; all features retained")
tse <- add_deviation(tse, "courses")
as.data.frame(recovery_results(tse, "courses"))[, c(
  "episode_id", "reference_support", "n_baseline_samples", "n_baseline_times", "baseline_diameter"
)]
```

References average **sample compositions**, not pooled counts. Each selected
sample has equal weight, including simultaneous samples. An explicit `features`
selection changes the compositional scope and must be justified before examining
outcomes. Do not compare distances computed on different feature sets as though
they measured the same quantity.

Support distinguishes `missing_baseline`, `single_sample`, `single_time` and
`multiple_times`. Diameter is the largest pairwise baseline Bray--Curtis distance;
it is missing with fewer than two samples. Replicates at one time do not establish
temporal stability. One baseline is usable but cannot measure baseline variation.
No baseline (`reference = character()`) is represented explicitly, without
borrowing another episode's profile. Support and diameter are descriptive, not
uncertainty intervals or automatically calibrated thresholds.

## State a rule before examining its outcome

Choose parameters for the biological question, measurement process and feasible
sampling design; record the rationale and plausible alternatives prospectively.
There are no universal settings:

| Parameter | Question to justify |
|:---|:---|
| `threshold` (0 to 1) | Which compositional departure is meaningful for this assay and feature scope? What baseline/technical variability can be assessed independently? |
| `persistence` (> 0) | Over what observed duration would return be relevant? Can the planned visits span it? |
| `max_gap` (> 0) | Which unobserved interval is tolerable between supporting visits? Is that assumption defensible? |
| `horizon` (> 0) | How long after the selected origin is the question defined, considering follow-up and other exposures? |

```{r rule}
rule <- list(threshold = 0.25, persistence = 4, max_gap = 3, horizon = 10)
tse <- add_recovery(tse, "courses", rule)
as.data.frame(recovery_results(tse, "courses"))[, c("episode_id", "status", "reason", "coverage")]
```

The sparse fixture illustrates the API, not a useful rule for that study. An
outside visit detects perturbation; a later within-band visit is a return.
Confirmation needs at least two distinct times spanning persistence, with no
outside visit between them and every supporting gap at most `max_gap`. Tied
samples count as one visit, which is within only if **all** are within. A rebound
is the first outside visit after confirmation, through the horizon. The gap from
perturbation to a candidate is not restricted by `max_gap`.

For an end origin, perturbation detection starts at exposure start (possibly
negative relative time); eligible returns start at zero. Boundaries are inclusive,
comparisons exact, and no interpolation or tolerance is silently added. Persistence
greater than the horizon is allowed but cannot yield confirmation.

`not_evaluable` reasons take precedence as missing baseline, overlapping events,
then no visits in the return window. A lack of return is not permanent non-return.
A single return is not continuous recovery. Visits after the horizon contribute
follow-up coverage only. Consult `?add_recovery` for the complete rule.

See the [real-data example](real-data.html) and its prespecified assessment for
baseline, sampling and rule sensitivity. To change a definition, register a new
analysis name and run its stages explicitly. There is no overwrite or implicit
refresh; an unsuccessful call leaves the input unchanged.

## Session information

```{r session-info}
utils::sessionInfo()
```
