---
title: "Personal baselines, observed recovery and analysis history"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Personal baselines, observed recovery and analysis history}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.alt = "Observed deviations and saved confirmation with incomplete current evidence."
)
```

## Current status

The experimental development version provides `setup_recovery()`,
`validate_recovery()`, `add_reference()`, `add_deviation()`, `add_recovery()`,
`recovery_results()` and `plot_recovery()`.
This vignette registers an analysis, attaches explicit personal references,
calculates sample deviations and observed episode outcomes, and checks their
dependencies while preserving historical records. It then extracts ordinary
result tables with their context and plots observations with saved time evidence.

## A small longitudinal input

The registration unit is an episode within a subject, with a perturbation
defined by events. Sample measurements and sample metadata remain in a
`TreeSummarizedExperiment` (TSE).

This deliberately small example has two features and three samples from one
subject. The time coordinate is days since enrolment, with an illustrative
exposure beginning on day 10. It demonstrates structure only; it cannot establish
sustained recovery or calibrate a statistical method. `recovery_examples`
contains this case, a second case with repeated episodes, and a third case for
observed recovery. All are synthetic; the package stores ordinary matrices and
tables and constructs a
TSE when needed.

```{r input-tse}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$single_episode

tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)

methods::validObject(tse)
SummarizedExperiment::assay(tse, "counts")
SummarizedExperiment::colData(tse)
```

A phylogenetic tree is not needed to construct this object. Any future method
that requires a tree will need to declare and validate that requirement.

## Declare episodes and events

An **episode** identifies the observation period relevant to one perturbation
and one subject. An **event** records a dated occurrence or interval, such as
the start and end of antibiotic exposure. Repeated perturbations need distinct
episode identities even when they occur in the same subject.

The episode selects the start or end of one of its events as its origin.
The tables below declare one episode and an exposure interval. Required
schemas and time rules are defined in the accepted
[RFC 001](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/001-registration-validation.md).

```{r registration-inputs}
episodes <- example_data$episodes
events <- example_data$events

episodes
events
```

Each sample is explicitly assigned to an episode by its `episode_id` column.
An `NA` episode explicitly excludes that sample from this named analysis;
each declared episode must still have an included sample. Registration accepts
finite numeric times with a declared unit and coordinate origin. Calendar
times need an explicit conversion by the caller before registration.

## Register and inspect the snapshot

`setup_recovery()` checks the identities, episode membership, and event
references, then returns the TSE with a named registration. Analysis IDs match
`^[a-z][a-z0-9]*$`. The corresponding `rec_<analysis>_` prefix is reserved for
future result columns; registration creates none.

```{r register-analysis}
tse <- recoverome::setup_recovery(
  tse,
  analysis_id = "antibiotic",
  episodes = episodes,
  events = events,
  time_col = example_data$time_col,
  time_unit = example_data$time_unit,
  time_origin = example_data$time_origin
)

registration <- S4Vectors::metadata(tse)$recoverome$analyses$antibiotic
registration$registration$samples
registration$episodes
registration$events
registration$scope
registration$owned_columns
```

The stored tables are `S4Vectors::DataFrame` objects. The sample snapshot
contains included samples in their original order; the scope records all
original sample and feature IDs, including any excluded samples. This is
historical metadata, and must not be edited as a second current data source.
Registration also records its source column names, the time declaration, the
package version, and a UTC registration timestamp.

Here the event origin is day 10, so the implied relative sample times are
`-7`, `0`, and `7` days. Registration retains the supplied coordinates without
creating a second time column. It neither selects an assay nor reads or
fingerprints its values.

Calling setup again with the same analysis ID is an error, even with identical
inputs. A pre-existing column beginning with `rec_antibiotic_` also blocks this
registration. Another unused name can register an independent analysis.

## Filtering preserves analysis history

Ordinary TSE subsetting selects features, samples, and their annotations while
preserving the original registration. This subset contains one feature and
only the sample collected after the exposure:

```{r subset-tse}
after_event <- tse["feature_a", "s3", drop = FALSE]
rownames(after_event)
colnames(after_event)
history <- S4Vectors::metadata(after_event)$recoverome$analyses$antibiotic
history$scope
identical(history, registration)
```

The stored scope still identifies both features and all three samples, and the
entire registration is unchanged. The snapshot describes the original data;
filtering does not make it a new registration of the subset.

`validate_recovery()` reports these distinctions without modifying the TSE.
Its return value is a list with `report_schema_version = 1L`, a `summary`
DataFrame, and a `diagnostics` DataFrame. Diagnostic IDs are a list-column.

```{r validate-history}
report <- recoverome::validate_recovery(after_event, analysis_id = "antibiotic")
report$summary
report$diagnostics
```

The record remains structurally valid and validation is complete. Both scopes
are reduced; the included-sample counts are three originally and one retained.
Dependencies are unchanged because the retained sample still matches the
recorded metadata. This does not certify the cells of removed samples.

Changing that retained sample's time is a separate finding:

```{r changed-metadata}
changed <- after_event
annotation <- SummarizedExperiment::colData(changed)
annotation[[example_data$time_col]] <- annotation[[example_data$time_col]] + 1
SummarizedExperiment::colData(changed) <- annotation
changed_report <- recoverome::validate_recovery(changed, analysis_id = "antibiotic")
changed_report$summary
changed_report$diagnostics
```

The historical snapshot remains structurally valid, while dependencies are
now changed. The diagnostic identifies `s3`; validation neither repairs the
metadata nor updates the stored snapshot. Registration validation does not
read assay values or establish their suitability for a later method.

## Repeated episodes and an excluded sample

The second case contains six samples and two episodes for the same subject.
Episode `e1` uses an event's start as its origin; `e2` uses its event's end.
The final sample has `NA` episode membership, so registration excludes it while
retaining it in the container and original scope.

```{r repeated-episodes}
repeated <- recovery_examples$repeated_episodes
repeated_tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = repeated$counts),
  colData = S4Vectors::DataFrame(repeated$col_data)
)
repeated_tse <- recoverome::setup_recovery(
  repeated_tse,
  analysis_id = "repeated",
  episodes = repeated$episodes,
  events = repeated$events,
  time_col = repeated$time_col,
  time_unit = repeated$time_unit,
  time_origin = repeated$time_origin
)
repeated_history <- S4Vectors::metadata(repeated_tse)$recoverome$analyses$repeated
repeated_history$registration$samples
repeated_history$scope$sample_ids
```

There are five registered rows and six original sample IDs. The corresponding
relative times are `-7`, `0`, `7`, `-8`, and `1` days. These arithmetic examples
illustrate explicit origins and membership, without defining recovery.

```{r excluded-only}
excluded_only <- repeated_tse[, "s6", drop = FALSE]
recoverome::validate_recovery(excluded_only, analysis_id = "repeated")$summary
```

Here none of the five registered samples remains, although the TSE still has
one column. The report keeps the historical episodes, reports zero retained
registered samples, and leaves dependencies `not_checked`. It does not invent
broken episodes or enrol the excluded sample.

## Attach an explicit personal baseline

`add_reference()` implements the reference contract in
[RFC 002](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/002-personal-baseline-deviation.md).
Give sample IDs rather than an inferred baseline window. Here `s1` belongs to
`e1` and `s4` to `e2`. Both precede their own origin event's start; `s4` at day 34
is eligible before start 40, even though `e2` selects end 42 as its origin.

```{r repeated-reference}
referenced <- recoverome::add_reference(
  repeated_tse,
  analysis_id = "repeated",
  reference = c("s1", "s4"),
  assay = "counts",
  preprocessing = "synthetic counts; no upstream transformations"
)
reference <- S4Vectors::metadata(referenced)$recoverome$analyses$repeated$reference
reference$baseline_samples
reference$profiles
reference$episodes
```

The episode profiles are `(1/3, 2/3)` and `(7/15, 8/15)`. Each has one selected
sample, so support is `single_sample` and its baseline diameter is `NA`. Multiple
selected samples would first be divided by their own selected-feature total,
then averaged with equal sample weight. A larger read total gives no extra
weight. Distinct samples at the same time count separately for this mean;
aggregate technical replicates explicitly upstream if needed.

Use `features` to declare a non-empty subcomposition, or leave it `NULL` to use
all currently retained features. Consumed values must be finite, non-negative
integer/double measurements with a finite positive sample total. Zeros are
allowed; no pseudocount, implicit feature filtering or missing-value deletion
is applied. Sparse and delayed matrix inputs are supported through extraction
of the requested block; its conversion can realize that block in memory.
The function makes no bound on backend I/O or memory use.

Reference selection never borrows samples from another episode. Passing
`reference = character()` records `missing_baseline` for every episode, without
zero-filled profiles. One baseline permits a descriptive reference but estimates
no variability. With multiple baselines, `baseline_diameter` is the maximum
pairwise Bray--Curtis dissimilarity of their compositions; it is not a recovery
cutoff or uncertainty interval. The function creates no deviation columns.

### Preserve the realized reference after filtering

The definition, baseline identities, profiles and input fingerprints remain
historical when baseline samples are later removed:

```{r reference-history}
reference_subset <- referenced[, c("s2", "s3", "s5"), drop = FALSE]
subset_reference <- S4Vectors::metadata(reference_subset)$recoverome$analyses$repeated$reference
identical(subset_reference, reference)
```

This retained record describes the original selected baselines. A later new
calculation that depends on those inputs requires them to remain available;
subsetting does not silently rebuild a reference. Use a new named analysis and
explicit selection to define a different reference. Repeating `add_reference()`
for the same analysis errors, even when the arguments are identical.

Validate the reference and its current baseline dependencies:

```{r reference-validation}
reference_report <- recoverome::validate_recovery(referenced, analysis_id = "repeated")
reference_report$summary
unique(reference_report$diagnostics$code)
```

The reference is structurally valid, validation is complete and dependencies
are unchanged. Its two single-sample baselines produce an informational support
finding, not an estimate of uncertainty. The validator compares the reference
fingerprint, registration parent and baseline measurements; it never estimates
the profiles again. Recovery records add their own checks when present;
unsupported schemas remain incompletely checked.

## Calculate deviations from the fixed references

`add_deviation()` uses the reference's recorded assay, features and profiles.
It checks the reference fingerprint, its registration dependency and selected
baseline inputs before creating a new result. Removing an unused feature or a
non-baseline sample is allowed; removing a selected feature or baseline sample
prevents a new calculation. Scaling baseline counts is also a source change,
even when the relative composition stays the same. Unrelated annotations and
assays do not redefine a dependency.

```{r repeated-deviations}
analysed <- recoverome::add_deviation(referenced, analysis_id = "repeated")
annotation <- SummarizedExperiment::colData(analysed)
annotation[, c("rec_repeated_deviation", "rec_repeated_deviation_status"), drop = FALSE]

# Independently stated values from the small input matrices in RFC 002.
expected <- c(0, 2 / 21, 4 / 33, 0, 2 / 285, NA_real_)
stopifnot(isTRUE(all.equal(annotation$rec_repeated_deviation, expected, tolerance = 1e-12)))
```

Both baselines have zero self-deviation here. Other included samples use their
own episode's profile, with no borrowing across episodes. Bray--Curtis lies in
[0, 1] for these compositions: zero means the selected proportions agree. It
is not a recovery threshold, health score or measure of absolute microbial load.
A one-feature reference gives constant compositions and zero deviations.

Only selected-feature values from retained included samples with a reference
are consumed. Invalid measurements in excluded samples or missing-baseline
episodes are not inspected or certified. Consumed data must remain finite,
non-negative and have a positive finite selected-feature total. Sparse and
delayed assays use the selected block; conversion can materialize that block.

The two result columns are authoritative. Metadata stores the method, their
exact name mapping, the current sample IDs as realized scope, the reference
fingerprint, per-computed-sample source hashes and per-sample output hashes.
These hashes document dependencies; they cannot reconstruct a removed value.

```{r deviation-provenance}
deviation <- S4Vectors::metadata(analysed)$recoverome$analyses$repeated$deviation
deviation$method
deviation$columns
deviation$sample_ids
```

An episode without a selected reference receives an explicit missing state:

```{r missing-baseline-deviation}
partial_reference <- recoverome::add_reference(
  repeated_tse,
  analysis_id = "repeated",
  reference = "s1",
  assay = "counts"
)
partial_analysis <- recoverome::add_deviation(partial_reference, analysis_id = "repeated")
SummarizedExperiment::colData(partial_analysis)[, c(
  "rec_repeated_deviation", "rec_repeated_deviation_status"
), drop = FALSE]
```

Here `s4` and `s5` are `missing_baseline`, and `s6` remains `excluded`. All three
have `NA` deviations with different reasons; invalid consumed measurements
would instead stop the call without modifying the input.

### Keep results historical after filtering

```{r deviation-history}
retained <- analysed[, c("s5", "s2"), drop = FALSE]
SummarizedExperiment::colData(retained)[, c(
  "rec_repeated_deviation", "rec_repeated_deviation_status"
), drop = FALSE]
retained_deviation <- S4Vectors::metadata(retained)$recoverome$analyses$repeated$deviation
identical(retained_deviation, deviation)
```

The retained values remain `2/285` and `2/21` in the new sample order; the
metadata still records all six IDs from the original calculation. Neither
filtering nor reordering recomputes the reference or the deviations. Calling
`add_deviation()` again errors, even with the same inputs. A changed definition
requires a new named analysis and explicit calculation of its stages.

## Validate sources, outputs and historical scope

`add_deviation()` checks dependencies when it calculates. `validate_recovery()`
checks the saved reference and deviations later without modifying the TSE.
The report retains the registration columns and adds analytical diagnostics;
no additional persistent result store or mutable validation flag is created.

```{r validate-deviations}
current_report <- recoverome::validate_recovery(analysed, analysis_id = "repeated")
current_report$summary

historical_report <- recoverome::validate_recovery(retained, analysis_id = "repeated")
historical_report$summary
historical_report$diagnostics
```

The full analysis has unchanged dependencies and complete validation. The
filtered analysis still has readable historical records, but the removed
baselines and samples prevent all original source/output comparisons. Its
validation is incomplete and dependencies are `not_checked`, rather than a
claim that its reference is unchanged for a new analysis of that subset.
Retained outputs are still checked where possible.

A current source edit is different from removal:

```{r changed-analytical-source}
source_changed <- analysed
SummarizedExperiment::assay(source_changed, "counts")[, "s2"] <- c(4L, 3L)
source_report <- recoverome::validate_recovery(source_changed, analysis_id = "repeated")
source_report$summary
source_report$diagnostics
```

This is a comparable input change, so dependencies are `changed` while
validation can remain complete. The stored reference and deviation structure
are intact. The original `s2` deviation remains `2/21`; the validator does not
replace it with a fresh calculation. Multiplying baseline counts would also
be a source change even if their proportions stayed identical.

An edited output remains detectable when source inputs are missing:

```{r missing-input-and-edited-output}
output_changed <- retained
SummarizedExperiment::colData(output_changed)$rec_repeated_deviation[1L] <- 0.5
before_validation <- serialize(output_changed, NULL)
output_report <- recoverome::validate_recovery(output_changed, analysis_id = "repeated")
output_report$summary
output_report$diagnostics
stopifnot(identical(serialize(output_changed, NULL), before_validation))
```

Here `s5` still has status `computed`, but its stored value no longer agrees
with the original output fingerprint. The report gives `structural_valid = FALSE`,
`dependencies = "changed"` and incomplete validation because other historical
inputs remain unavailable. Both findings matter: missing inputs do not hide
the independently detectable output change.

In general, known changes take precedence over unavailable comparisons in the
`dependencies` summary. A comparison that detects a difference can be complete;
a missing or unreadable required input cannot. Unknown schemas or fingerprint
formats remain incomplete, and an unknown encoding is not treated as proof
that values changed. Other interpretable stages and analyses are still checked.

Checks use recorded identities, consumed assay blocks, parent records and owned
outputs. Unrelated annotations and assays are not dependencies. Registration-only
analyses still need no assay access. Source/result hashes detect ordinary edits;
manual rewriting of stored snapshots is unsupported. No validation state proves
biological recovery. See the
[analytical diagnostic guide](https://github.com/xec-cm/recoverome/blob/devel/dev/analytical-validation.md)
for the code catalogue and detection boundaries.

## Observed recovery by episode

The implemented rule follows [RFC 003](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/003-observed-recovery.md).
Every parameter is explicit; this example uses a threshold of 0.25, persistence
of 4 days, maximum consecutive gap of 3 days and horizon of 10 days.
These choices illustrate the method rather than establish biological limits.

```{r observed-recovery}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$observed_recovery
tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
tse <- recoverome::setup_recovery(
  tse,
  analysis_id = "observed",
  episodes = example_data$episodes,
  events = example_data$events,
  time_col = example_data$time_col,
  time_unit = example_data$time_unit,
  time_origin = example_data$time_origin
)
tse <- recoverome::add_reference(tse, "observed", reference = "b1", assay = "counts")
tse <- recoverome::add_deviation(tse, "observed")
rule <- list(threshold = 0.25, persistence = 4, max_gap = 3, horizon = 10)
recovered <- recoverome::add_recovery(tse, "observed", rule)
outcome <- S4Vectors::metadata(recovered)$recoverome$analyses$observed$recovery
outcome$episodes[, c("status", "candidate_time", "confirmation_time", "rebound_time", "coverage")]
```

Here the observed deviation at day 0 is 0.75. Visits at days 2, 4 and 6 are
within the band and span four days, with gaps no greater than three days.
The candidate is day 2 and confirmation is day 6. The outside observation at
day 8 is a rebound; it does not erase the first confirmation.

```{r observed-evidence}
outcome$evidence$episode_1
recoverome::validate_recovery(recovered, "observed")$summary
```

Visits at the same exact time are grouped conservatively: every sample must be
within the band. A larger gap or outside visit breaks a run. Observations after
the horizon contribute to coverage only, not to detection, confirmation or
rebound. Other registered events of the same subject overlapping the episode's
analysis interval make it not evaluable, even when they occur after a possible
confirmation. Missing baseline takes precedence over overlapping events, then
absence of observations in the return window.

`status` distinguishes `not_evaluable`, `no_detected_perturbation`,
`no_observed_return`, `unconfirmed_return` and `confirmed_return`. For
`not_evaluable`, `reason` records `missing_baseline`, `unresolved_events` or
`no_observations_in_window`. Coverage is separately `none`,
`ends_before_horizon` or `reaches_horizon`. A confirmed return need not have
follow-up through the entire horizon.

Filtering preserves historical confirmation and its evidence:

```{r filtered-recovery}
filtered_recovery <- recovered[, colnames(recovered) != "s3"]
before_validation <- serialize(filtered_recovery, NULL)
filtered_report <- recoverome::validate_recovery(filtered_recovery, "observed")
filtered_report$summary
stopifnot(
  identical(serialize(filtered_recovery, NULL), before_validation),
  identical(
    S4Vectors::metadata(filtered_recovery)$recoverome$analyses$observed$recovery,
    outcome
  )
)
```

Day-4 sample `s3` supported confirmation. Its removal leaves the historical
outcome intact, with incomplete validation and `not_checked` dependencies.
It does not turn the episode into a newly evaluated unconfirmed return.
Changing an available parent or output is diagnosed independently. Read the
[implementation guide](https://github.com/xec-cm/recoverome/blob/devel/dev/observed-recovery.md)
for storage and validation boundaries.

## Extract ordinary result tables

`recovery_results()` selects one analysis and validates it once, without
recomputing its saved values. Both levels return an `S4Vectors::DataFrame`
with atomic identity, availability and validation columns.

```{r episode-results}
episodes <- recoverome::recovery_results(filtered_recovery, "observed")
episodes[, c("episode_id", "result_state", "confirmation_time", "dependencies")]
stopifnot(
  identical(episodes$confirmation_time, 6),
  identical(episodes$result_state, "available"),
  identical(episodes$validation_complete, FALSE)
)
```

The authoritative episode outcome remains available even when some supporting
samples are absent. Its flags describe the selected analysis as a whole.
At sample level, removed values cannot be reconstructed from fingerprints:

```{r sample-results}
samples <- recoverome::recovery_results(
  filtered_recovery, "observed", level = "sample", scope = "historical"
)
as.data.frame(samples)[, c("sample_id", "relative_time", "result_state", "deviation")]
removed <- samples[samples$sample_id == "s3", ]
stopifnot(
  identical(removed$relative_time, 4),
  identical(removed$result_state, "removed"),
  identical(removed$deviation, NA_real_),
  identical(removed$deviation_status, NA_character_)
)
```

`scope = "current"` is the default and selects retained original sample IDs,
or episodes with retained included samples. `"historical"` selects all
original IDs. Both follow registration order. Newly added IDs produce one
structured warning; they are never enrolled. Excluded samples have no invented
subject, episode or time. Missing stages use typed NA and `not_computed`;
a stored `missing_baseline` deviation remains an available result.

Saved coordinates are used even after current source annotations change:

```{r saved-result-coordinates}
changed_times <- recovered
SummarizedExperiment::colData(changed_times)$day[5] <- 30
saved <- recoverome::recovery_results(changed_times, "observed", level = "sample")
stopifnot(
  identical(saved$relative_time[saved$sample_id == "s4"], 6),
  all(saved$dependencies == "changed")
)
```

Changed or unavailable current inputs leave readable saved results accessible,
with their diagnostic severities preserved. Broken identities, inconsistent
saved outputs or parent records, and required unsupported schemas error.
The accessor never repairs or refits an analysis.

The full context remains outside the atomic columns:

```{r result-context}
context <- S4Vectors::metadata(episodes)$recoverome_view
context$validation$summary
context$evidence$episode_1$confirmation_run
stopifnot(identical(context$evidence$episode_1$confirmation_run, c("s2", "s3", "s4")))
```

The context includes registration/events, original and realized scopes,
definitions, existing provenance and fingerprints, and full evidence even for
removed samples. It is an extraction-time snapshot. Filtering or editing the
returned table does not refresh it. Call the accessor again against the TSE
for a fresh report. `as.data.frame()` preserves the atomic columns;
`tibble::as_tibble(as.data.frame(samples))` is an optional downstream conversion
with no promise to preserve metadata. Optional tidy operations on the TSE
itself are a separate development issue.

## Plotting observed evidence

```{r observed-plot, fig.width=9, fig.height=7}
p <- recoverome::plot_recovery(filtered_recovery, "observed", scope = "historical")
p + ggplot2::labs(title = "Historical confirmation after filtering")
```

`plot_recovery()` returns an ordinary unprinted ggplot object. Customize it with
`+ ggplot2::labs(...)` or `+ ggplot2::theme(...)`. `ggplot2` supplies the layers and
is a runtime dependency. A deviation stage is required. Before `add_recovery()`,
the plot shows deviations neutrally with no invented threshold or outcome.

Facets identify the subject, episode, original baseline support and current
baseline count, selected origin event/boundary, and saved status, reason and
coverage. `episodes` accepts an ordered vector of IDs within the selected scope.
All facets share the deviation scale [0, 1] and the original included time
extent, zero and any declared horizon. Points beyond the horizon remain outside
the rule window. Edited current annotations cannot move saved observations.

The separate evidence rail shows saved milestone times. A candidate-to-confirmation
bracket represents observed supporting span, not continuous recovery. Hollow marks
and dashed spans indicate missing current evidence IDs, not uncertain timing.
Confirmation needs the entire saved run, so removing an intermediate observation
leaves its retained endpoint point solid but its confirmation marker hollow.
Removed times are crosses on the rail, never invented deviations.

Original evaluated gaps greater than the rule's maximum gap remain visible;
filtering cannot create a new original gap or re-evaluate a run. Yellow shading
marks the region beyond saved follow-up when coverage ends before the horizon.
Missing follow-up is labeled separately. Neither region establishes an outside
state. The declared detection-window shading is not a confidence interval or
an estimate derived from baseline diameter.

The plot states current or historical scope. Source changes and incomplete
checks appear in its caption and each facet. Historical episodes with no
remaining values retain their saved evidence rail. The TSE and its historical
outcomes are preserved; customizing the plot does not revalidate it.

## Interpretation boundaries

Observed similarity at one visit does not establish a sustained return. A rule
based on repeated visits must also state the relevant duration and admissible
gaps. Sparse observations cannot establish continuous behavior between visits
without additional assumptions.

Likewise, the absence of an observed recovery by the end of follow-up does not
establish that recovery will never occur. Methods for uncertainty, incomplete
follow-up, and comparisons between groups will need dedicated statistical
validation before implementation claims can be made.

The intended outcome is an explicit, auditable description of recovery under
a stated definition. Recovery to baseline is not a clinical diagnosis or a
claim that all microbial functions have returned.

## Session information

```{r session-info}
sessionInfo()
```
