---
title: "Get started: from samples to observed return"
author: "Francesc Catala-Moll"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Get started: from samples to observed return}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(recoverome)
```

## Introduction

This five-minute workflow describes return to a personal microbiome reference,
extracts an interpretable table and makes a first plot. It runs offline with a
small synthetic dataset. The values illustrate the calculation; they are **not
recommended biological cutoffs**. Similarity to a baseline is not proof of health,
functional restoration or continuous recovery between visits.

This workflow complements general microbiome processing with
[mia](https://bioconductor.org/packages/mia/) and ecological dissimilarities in
[vegan](https://vegandevs.github.io/vegan/reference/vegdist.html) by retaining
explicit longitudinal definitions and historical outcomes in the TSE.

## Installation

The experimental `devel` branch requires R >= 4.6.0. Version 0.99.0 is a submission
candidate, not an accepted Bioconductor release. Install from GitHub:

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("xec-cm/recoverome", ref = "devel")
```

## Run a complete analysis

The example has one participant, two features and seven samples. `b1` is the
explicit baseline, collected before an exposure running from day 10 to day 14.
Time zero is the exposure's start. The TSE holds the abundance matrix and sample
annotations; the two supplied tables describe the episode and its exposure.

Each sample is converted to a composition. The personal reference averages the
selected baseline compositions with equal sample weights. `add_deviation()`
calculates Bray--Curtis against that fixed reference.

```{r complete-analysis}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$observed_recovery
raw <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
tse <- setup_recovery(
  raw,
  analysis_id = "observed",
  episodes = example_data$episodes,
  events = example_data$events,
  time_col = "day",
  time_unit = "days",
  time_origin = "days since enrolment"
)
tse <- add_reference(tse, "observed", reference = "b1", assay = "counts",
                     preprocessing = "synthetic counts; no upstream transformations")
tse <- add_deviation(tse, "observed")
rule <- list(threshold = 0.25, persistence = 4, max_gap = 3, horizon = 10)
tse <- add_recovery(tse, "observed", rule)

```

The four rule values mean: deviation at most **0.25**, observations spanning at
least **4 days**, consecutive supporting visits at most **3 days apart**, and a
**10-day horizon**. They must be specified explicitly for your own study.

## Read the result

```{r episode-result}
episodes <- recovery_results(tse, "observed")
as.data.frame(episodes)[, c(
  "episode_id", "status", "candidate_time", "confirmation_time", "rebound_time", "coverage"
)]
```

The first eligible return is day **2**. Visits at days 2, 4 and 6 confirm a run
at day **6**; an outside visit at day **8** records a rebound. Confirmation and
rebound can coexist: the rule describes observations, not permanent recovery.
`reaches_horizon` means the last observation reaches the declared horizon; it
does not guarantee adequate sampling between visits.

```{r plot, fig.width=8, fig.height=6, fig.alt="Synthetic deviations with return at day 2, confirmation at day 6 and rebound at day 8."}
plot_recovery(tse, "observed")
```

The plot is an editable ggplot. Points show observed deviations; the evidence
rail records the saved milestones. Shading represents the rule, not a confidence
interval. Large gaps are not evidence of what happened between measurements.

## Inspect samples and check the analysis

```{r samples-and-validation}
samples <- recovery_results(tse, "observed", level = "sample")
as.data.frame(samples)[, c("sample_id", "relative_time", "deviation", "deviation_status")]
validate_recovery(tse, "observed")$summary
stopifnot(
  identical(samples$deviation, c(0, .75, .25, .125, .125, .5, .125)),
  episodes$confirmation_time == 6, episodes$rebound_time == 8
)
```

The reference is `(1, 0)`, so deviation equals the second feature's proportion.
The last assertion checks these independently stated numbers. Validation reports
whether stored records and available inputs agree; it does not assess biological
validity. All operations above preserve assays and unrelated TSE annotations.

## Continue with your own data

- [Prepare inputs](input-preparation.html): identities, episodes, events, dates,
  baseline choice and explicit rules.
- [Filtering, history and validation](history-and-validation.html): what changes
  when samples disappear and when a new analysis is needed.
- [A real antibiotic-exposure example](real-data.html): original observations,
  baseline support, sparse follow-up and the scientific assessment.

Use a new analysis name to change a definition. Setup/add functions do not
overwrite a stage; filtering and validation never silently recompute it.

## Session information

```{r session-info}
utils::sessionInfo()
```
