---
title: "Recovery analysis: data structure and planned workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recovery analysis: data structure and planned workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Current status

`recoverome` 0.1.0 is an experimental scaffold. Recovery functions are not
implemented yet. This vignette constructs a valid input object and explains
the intended contracts. It does not calculate recovery outcomes.

## A small longitudinal input

The planned analysis unit is an episode within a subject, with a perturbation
defined by events. Sample measurements and sample metadata remain in a
`TreeSummarizedExperiment` (TSE).

This deliberately small example has two features and three samples from one
subject. The time coordinate is days relative to an illustrative event at
day zero. It demonstrates structure only; it is not sufficient to establish
sustained recovery or to calibrate a statistical method.

```{r input-tse}
counts <- matrix(
  c(80L, 20L, 75L, 25L, 40L, 60L),
  nrow = 2L,
  dimnames = list(c("feature_a", "feature_b"), c("s1", "s2", "s3"))
)

sample_data <- S4Vectors::DataFrame(
  subject_id = rep("participant_1", 3L),
  day = c(-7, -1, 7),
  phase = c("before", "before", "after"),
  row.names = colnames(counts)
)

tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = counts),
  colData = sample_data
)

methods::validObject(tse)
SummarizedExperiment::assay(tse, "counts")
SummarizedExperiment::colData(tse)
```

A phylogenetic tree is not needed to construct this object. Any future method
that requires a tree will need to declare and validate that requirement.

## Episodes, events, and references

An **episode** identifies the observation period relevant to one perturbation
and one subject. An **event** records a dated occurrence or interval, such as
the start and end of antibiotic exposure. Repeated perturbations need distinct
episode identities even when they occur in the same subject.

A **reference** records what return means for a named analysis. It may be based
on selected pre-event samples, but the exact eligibility rules and the samples
actually used must remain visible. A reference profile and a recovery margin
are different concepts: the former defines the comparison target, while the
latter defines an acceptable deviation under the chosen rule.

The initial contracts will keep these records in
`S4Vectors::metadata(tse)$recoverome$analyses[[analysis_id]]`. Sample-level
annotations will be placed in `colData()` columns with the prefix
`rec_<analysis>_`; for example, `rec_antibiotic_deviation`. Analysis IDs will be
simple, stable identifiers validated to prevent prefix collisions. Detailed
record schemas will be settled with their implementation.

## Proposed analysis interface

The following block sketches the planned workflow. **It is not executable with
this version of recoverome.** Objects such as `episodes`, `events`, `reference`,
and `recovery_rule` denote future validated inputs; their schemas are not
presented here as implemented interfaces.

```{r planned-workflow, eval=FALSE}
tse <- setup_recovery(
  tse,
  analysis_id = "antibiotic",
  episodes = episodes,
  events = events
)
tse <- add_reference(tse, analysis_id = "antibiotic", reference = reference)
tse <- add_deviation(tse, analysis_id = "antibiotic")
tse <- add_recovery(tse, analysis_id = "antibiotic", rule = recovery_rule)

validate_recovery(tse, analysis_id = "antibiotic")
recovery_results(tse, analysis_id = "antibiotic", level = "episode")
plot_recovery(tse, analysis_id = "antibiotic")
```

The four setup/add functions are intended to return an updated TSE. Extraction,
plotting, and validation have separate roles and must not silently alter an
analysis. Statistical fitting is a future layer outside this initial interface.

## Filtering preserves analysis history

Ordinary TSE subsetting already selects samples and their sample annotations:

```{r subset-tse}
after_event <- tse[, SummarizedExperiment::colData(tse)$day > 0]
colnames(after_event)
SummarizedExperiment::colData(after_event)
```

This example has no recoverome analysis attached. In the planned contract,
subsetting an annotated TSE will preserve analysis history and its original
sample and feature scope. Removing a reference sample or a follow-up visit
must not cause an automatic refit or a new recovery time.

The planned validator will identify mismatches between the present data and
the recorded analysis scope. A result computed from the original data remains
a historical result; it must not be presented as a new analysis of the subset.
Intentional recomputation will require an explicit analysis operation.

## Interpretation boundaries

Observed similarity at one visit does not establish a sustained return. A rule
based on repeated visits must also state the relevant duration and admissible
gaps. Sparse observations cannot establish continuous behavior between visits
without additional assumptions.

Likewise, the absence of an observed recovery by the end of follow-up does not
establish that recovery will never occur. Methods for uncertainty, incomplete
follow-up, and comparisons between groups will need dedicated statistical
validation before implementation claims can be made.

The intended outcome is an explicit, auditable description of recovery under
a stated definition. Recovery to baseline is not a clinical diagnosis or a
claim that all microbial functions have returned.

## Session information

```{r session-info}
sessionInfo()
```
