---
title: "Register an analysis and preserve its history"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Register an analysis and preserve its history}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Current status

The experimental development version provides `setup_recovery()`,
`validate_recovery()` and `add_reference()`. This vignette registers an analysis,
diagnoses its registration history, and attaches an explicit personal reference.
Deviation and recovery calculations, extraction, and plotting remain planned.

## A small longitudinal input

The registration unit is an episode within a subject, with a perturbation
defined by events. Sample measurements and sample metadata remain in a
`TreeSummarizedExperiment` (TSE).

This deliberately small example has two features and three samples from one
subject. The time coordinate is days since enrolment, with an illustrative
exposure beginning on day 10. It demonstrates structure only; it cannot establish
sustained recovery or calibrate a statistical method. `recovery_examples`
contains this case and a second case with repeated episodes. Both are fully
synthetic; the package stores ordinary matrices and tables and constructs a
TSE when needed.

```{r input-tse}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$single_episode

tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)

methods::validObject(tse)
SummarizedExperiment::assay(tse, "counts")
SummarizedExperiment::colData(tse)
```

A phylogenetic tree is not needed to construct this object. Any future method
that requires a tree will need to declare and validate that requirement.

## Declare episodes and events

An **episode** identifies the observation period relevant to one perturbation
and one subject. An **event** records a dated occurrence or interval, such as
the start and end of antibiotic exposure. Repeated perturbations need distinct
episode identities even when they occur in the same subject.

The episode selects the start or end of one of its events as its origin.
The tables below declare one episode and an exposure interval. Required
schemas and time rules are defined in the accepted
[RFC 001](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/001-registration-validation.md).

```{r registration-inputs}
episodes <- example_data$episodes
events <- example_data$events

episodes
events
```

Each sample is explicitly assigned to an episode by its `episode_id` column.
An `NA` episode explicitly excludes that sample from this named analysis;
each declared episode must still have an included sample. Registration accepts
finite numeric times with a declared unit and coordinate origin. Calendar
times need an explicit conversion by the caller before registration.

## Register and inspect the snapshot

`setup_recovery()` checks the identities, episode membership, and event
references, then returns the TSE with a named registration. Analysis IDs match
`^[a-z][a-z0-9]*$`. The corresponding `rec_<analysis>_` prefix is reserved for
future result columns; registration creates none.

```{r register-analysis}
tse <- recoverome::setup_recovery(
  tse,
  analysis_id = "antibiotic",
  episodes = episodes,
  events = events,
  time_col = example_data$time_col,
  time_unit = example_data$time_unit,
  time_origin = example_data$time_origin
)

registration <- S4Vectors::metadata(tse)$recoverome$analyses$antibiotic
registration$registration$samples
registration$episodes
registration$events
registration$scope
registration$owned_columns
```

The stored tables are `S4Vectors::DataFrame` objects. The sample snapshot
contains included samples in their original order; the scope records all
original sample and feature IDs, including any excluded samples. This is
historical metadata, and must not be edited as a second current data source.
Registration also records its source column names, the time declaration, the
package version, and a UTC registration timestamp.

Here the event origin is day 10, so the implied relative sample times are
`-7`, `0`, and `7` days. Registration retains the supplied coordinates without
creating a second time column. It neither selects an assay nor reads or
fingerprints its values.

Calling setup again with the same analysis ID is an error, even with identical
inputs. A pre-existing column beginning with `rec_antibiotic_` also blocks this
registration. Another unused name can register an independent analysis.

## Filtering preserves analysis history

Ordinary TSE subsetting selects features, samples, and their annotations while
preserving the original registration. This subset contains one feature and
only the sample collected after the exposure:

```{r subset-tse}
after_event <- tse["feature_a", "s3", drop = FALSE]
rownames(after_event)
colnames(after_event)
history <- S4Vectors::metadata(after_event)$recoverome$analyses$antibiotic
history$scope
identical(history, registration)
```

The stored scope still identifies both features and all three samples, and the
entire registration is unchanged. The snapshot describes the original data;
filtering does not make it a new registration of the subset.

`validate_recovery()` reports these distinctions without modifying the TSE.
Its return value is a list with `report_schema_version = 1L`, a `summary`
DataFrame, and a `diagnostics` DataFrame. Diagnostic IDs are a list-column.

```{r validate-history}
report <- recoverome::validate_recovery(after_event, analysis_id = "antibiotic")
report$summary
report$diagnostics
```

The record remains structurally valid and validation is complete. Both scopes
are reduced; the included-sample counts are three originally and one retained.
Dependencies are unchanged because the retained sample still matches the
recorded metadata. This does not certify the cells of removed samples.

Changing that retained sample's time is a separate finding:

```{r changed-metadata}
changed <- after_event
annotation <- SummarizedExperiment::colData(changed)
annotation[[example_data$time_col]] <- annotation[[example_data$time_col]] + 1
SummarizedExperiment::colData(changed) <- annotation
changed_report <- recoverome::validate_recovery(changed, analysis_id = "antibiotic")
changed_report$summary
changed_report$diagnostics
```

The historical snapshot remains structurally valid, while dependencies are
now changed. The diagnostic identifies `s3`; validation neither repairs the
metadata nor updates the stored snapshot. Registration validation does not
read assay values or establish their suitability for a later method.

## Repeated episodes and an excluded sample

The second case contains six samples and two episodes for the same subject.
Episode `e1` uses an event's start as its origin; `e2` uses its event's end.
The final sample has `NA` episode membership, so registration excludes it while
retaining it in the container and original scope.

```{r repeated-episodes}
repeated <- recovery_examples$repeated_episodes
repeated_tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = repeated$counts),
  colData = S4Vectors::DataFrame(repeated$col_data)
)
repeated_tse <- recoverome::setup_recovery(
  repeated_tse,
  analysis_id = "repeated",
  episodes = repeated$episodes,
  events = repeated$events,
  time_col = repeated$time_col,
  time_unit = repeated$time_unit,
  time_origin = repeated$time_origin
)
repeated_history <- S4Vectors::metadata(repeated_tse)$recoverome$analyses$repeated
repeated_history$registration$samples
repeated_history$scope$sample_ids
```

There are five registered rows and six original sample IDs. The corresponding
relative times are `-7`, `0`, `7`, `-8`, and `1` days. These arithmetic examples
illustrate explicit origins and membership, without defining recovery.

```{r excluded-only}
excluded_only <- repeated_tse[, "s6", drop = FALSE]
recoverome::validate_recovery(excluded_only, analysis_id = "repeated")$summary
```

Here none of the five registered samples remains, although the TSE still has
one column. The report keeps the historical episodes, reports zero retained
registered samples, and leaves dependencies `not_checked`. It does not invent
broken episodes or enrol the excluded sample.

## Attach an explicit personal baseline

`add_reference()` implements the reference contract in
[RFC 002](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/002-personal-baseline-deviation.md).
Give sample IDs rather than an inferred baseline window. Here `s1` belongs to
`e1` and `s4` to `e2`. Both precede their own origin event's start; `s4` at day 34
is eligible before start 40, even though `e2` selects end 42 as its origin.

```{r repeated-reference}
referenced <- recoverome::add_reference(
  repeated_tse,
  analysis_id = "repeated",
  reference = c("s1", "s4"),
  assay = "counts",
  preprocessing = "synthetic counts; no upstream transformations"
)
reference <- S4Vectors::metadata(referenced)$recoverome$analyses$repeated$reference
reference$baseline_samples
reference$profiles
reference$episodes
```

The episode profiles are `(1/3, 2/3)` and `(7/15, 8/15)`. Each has one selected
sample, so support is `single_sample` and its baseline diameter is `NA`. Multiple
selected samples would first be divided by their own selected-feature total,
then averaged with equal sample weight. A larger read total gives no extra
weight. Distinct samples at the same time count separately for this mean;
aggregate technical replicates explicitly upstream if needed.

Use `features` to declare a non-empty subcomposition, or leave it `NULL` to use
all currently retained features. Consumed values must be finite, non-negative
integer/double measurements with a finite positive sample total. Zeros are
allowed; no pseudocount, implicit feature filtering or missing-value deletion
is applied. Sparse and delayed matrix inputs are supported through extraction
of the requested block; its conversion can realize that block in memory.
The function makes no bound on backend I/O or memory use.

Reference selection never borrows samples from another episode. Passing
`reference = character()` records `missing_baseline` for every episode, without
zero-filled profiles. One baseline permits a descriptive reference but estimates
no variability. With multiple baselines, `baseline_diameter` is the maximum
pairwise Bray--Curtis dissimilarity of their compositions; it is not a recovery
cutoff or uncertainty interval. The function creates no deviation columns.

### Preserve the realized reference after filtering

The definition, baseline identities, profiles and input fingerprints remain
historical when baseline samples are later removed:

```{r reference-history}
reference_subset <- referenced[, c("s2", "s3", "s5"), drop = FALSE]
subset_reference <- S4Vectors::metadata(reference_subset)$recoverome$analyses$repeated$reference
identical(subset_reference, reference)
```

This retained record describes the original selected baselines. A later new
calculation that depends on those inputs requires them to remain available;
subsetting does not silently rebuild a reference. Use a new named analysis and
explicit selection to define a different reference. Repeating `add_reference()`
for the same analysis errors, even when the arguments are identical.

The registration validator has not yet been extended to analytical stages:

```{r reference-validation-boundary}
reference_report <- recoverome::validate_recovery(referenced, analysis_id = "repeated")
reference_report$summary
unique(reference_report$diagnostics$code)
```

`STAGE_UNSUPPORTED` and incomplete validation are expected here. Registration
findings can still be reported, but this diagnostic does not yet verify reference
fingerprints or assay changes. `add_reference()` validates the inputs it consumes
when building the reference. Subsequent stage-aware dependency validation belongs
to issue #12; a created reference must not be presented as fully validated by
this registration-only report.

## Planned analytical stages

A recovery rule will define an acceptable deviation and observational requirements.
The following is **non-executable proposed API**; these four functions are not
implemented. Their accepted design contracts remain distinct from available code.

```{r planned-workflow, eval=FALSE}
analysed <- add_deviation(referenced, analysis_id = "repeated")
analysed <- add_recovery(analysed, analysis_id = "repeated", rule = recovery_rule)

recovery_results(analysed, analysis_id = "repeated", level = "episode")
plot_recovery(analysed, analysis_id = "repeated")
```

The add functions are intended to return a TSE. Extraction, plotting, and
validation have separate roles and must not silently alter an analysis.
Statistical fitting is a future layer outside this initial interface.

## Interpretation boundaries

Observed similarity at one visit does not establish a sustained return. A rule
based on repeated visits must also state the relevant duration and admissible
gaps. Sparse observations cannot establish continuous behavior between visits
without additional assumptions.

Likewise, the absence of an observed recovery by the end of follow-up does not
establish that recovery will never occur. Methods for uncertainty, incomplete
follow-up, and comparisons between groups will need dedicated statistical
validation before implementation claims can be made.

The intended outcome is an explicit, auditable description of recovery under
a stated definition. Recovery to baseline is not a clinical diagnosis or a
claim that all microbial functions have returned.

## Session information

```{r session-info}
sessionInfo()
```
