---
title: "A complete personal-baseline recovery workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A complete personal-baseline recovery workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this workflow answers

Given explicitly selected personal baseline samples and a known perturbation,
when do the observed compositions return within a declared band, and do later
visits support confirmation under a declared persistence and gap rule?

This vignette uses all seven implemented functions in the experimental version
of recoverome. Run it with the installed package in a fresh R session; it needs
no downloaded dataset, source-tree helpers or tidy adapter. The bundled data
are synthetic and the rule is illustrative. Neither the example nor its
confirmation establishes clinical recovery, baseline stability or continuous
behavior between visits.

The [README](https://xec-cm.github.io/recoverome/) gives a shorter quickstart.
Here we inspect the input, verify independently stated numerical results, and
follow the same completed analysis through extraction, plotting and filtering.

## 1. Build the input and register its meaning

Our example contains two features and seven samples from one participant.
Samples are TSE columns, features are rows, and both have explicit unique IDs.
Sample annotations supply subject, episode and numeric time. An **episode**
identifies the observations associated with one perturbation in one subject;
an **event** records a dated occurrence or interval. An episode selects the
start or end of its origin event as relative time zero.

```{r input-tse}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$observed_recovery

tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
SummarizedExperiment::assay(tse, "counts")
SummarizedExperiment::colData(tse)
example_data$episodes
example_data$events
```

These times are days since enrolment: `8, 10, 12, 14, 16, 18, 20`. The exposure
starts on day 10 and ends on day 14; this episode chooses its **start** as zero.
Thus relative sample times are `-2, 0, 2, 4, 6, 8, 10` days. The time coordinate
description and the episode's chosen event boundary have different roles.
No phylogenetic tree is required by the implemented compositional method;
existing trees and their links remain part of an input TSE.

```{r register-analysis}
tse <- recoverome::setup_recovery(
  tse,
  analysis_id = "observed",
  episodes = example_data$episodes,
  events = example_data$events,
  time_col = example_data$time_col,
  time_unit = example_data$time_unit,
  time_origin = example_data$time_origin
)
```

The setup/add functions return the TSE. Registration stores its input snapshot,
original scope and provenance in
`S4Vectors::metadata(tse)$recoverome$analyses$observed`; it does not read assay
values or create sample results. Do not edit that historical snapshot as a
second current data source. Analysis names match `^[a-z][a-z0-9]*$`; the name
`observed` reserves the `rec_observed_` sample-column prefix. Repeating setup or
an add stage errors rather than overwriting it. Use a new name and explicit
calculations for a different analysis.

A start origin permits confirmation during the exposure. Choose an end origin
at registration if the question requires post-exposure return; that choice
changes the relative coordinates and is not inferred by the package.

The accepted [registration contract][rfc1] defines required table schemas,
identities and time declarations. Calendar times require explicit conversion
to numeric coordinates by the caller. An `NA` episode annotation excludes a
sample from the analysis while preserving it in the container and original
scope; every declared episode must still have an included sample.

## 2. Attach an explicit personal reference

Select baseline sample IDs explicitly. Here `b1` at day 8 precedes the origin
event's start. Selection must precede that **start**, even for an episode that
uses the event's end as its time origin. No baseline window is inferred and no
sample is borrowed from a different episode.

```{r personal-reference}
tse <- recoverome::add_reference(
  tse,
  analysis_id = "observed",
  reference = "b1",
  assay = "counts",
  preprocessing = "synthetic counts; no upstream transformations"
)
reference <- S4Vectors::metadata(tse)$recoverome$analyses$observed$reference
reference$baseline_samples
reference$profiles
reference$episodes
```

Counts `(8, 0)` give composition `(1, 0)`. With multiple selected baselines,
each sample is divided by its own selected-feature total, then compositions
are averaged with equal sample weight. Larger read totals do not give more
weight. The [baseline and deviation contract][rfc2] fixes this definition.

One baseline yields `single_sample` support and `NA` baseline diameter: it
cannot estimate temporal variability. With several baselines the diameter is
the maximum pairwise Bray--Curtis dissimilarity of their compositions. It is
not a recovery cutoff or uncertainty interval. Missing baseline support is
recorded explicitly, with no fabricated zero profile.

The default uses all currently retained features; `features` can declare a
non-empty subcomposition. Consumed values must be finite and non-negative,
with positive finite selected-feature totals. Zeros are allowed. No pseudocount,
implicit filtering or missing-value deletion is applied. Sparse and delayed
inputs are supported through selected-block extraction, which can realize that
block in memory. Record relevant upstream preprocessing in your analysis.

## 3. Calculate and check sample deviations

`add_deviation()` uses the reference's recorded assay, features and fixed
profiles. It checks its required reference dependencies before calculating.
Complete these analytical stages before removing required inputs.

```{r sample-deviations}
tse <- recoverome::add_deviation(tse, "observed")
sample_results <- recoverome::recovery_results(tse, "observed", level = "sample")
as.data.frame(sample_results)[, c(
  "sample_id", "relative_time", "is_reference", "deviation", "deviation_status"
)]
```

For unit-sum compositions, Bray--Curtis is half the sum of absolute component
differences. Relative to `(1, 0)`, composition `(1 - d, d)` therefore has
dissimilarity `d`. The seven count pairs independently give the values below:

```{r verify-deviations}
expected_deviation <- c(0, 3 / 4, 1 / 4, 1 / 8, 1 / 8, 1 / 2, 1 / 8)
stopifnot(isTRUE(all.equal(
  sample_results$deviation, expected_deviation, tolerance = 1e-12
)))
```

The tolerance allows floating-point arithmetic; it does not represent
measurement uncertainty. Dissimilarity lies in `[0, 1]` and describes selected
proportions, not absolute microbial load or health. Zero self-deviation for this
single baseline is expected, not evidence of stable baseline variation.

The authoritative sample columns are `rec_observed_deviation` and
`rec_observed_deviation_status` in `colData(tse)`. Metadata records their method,
identities, dependencies and provenance. Status `computed` means a past
calculation exists; it does not assert present validity. Missing-baseline
samples receive `NA` with `missing_baseline`; excluded samples receive `NA`
with `excluded`. Unconsumed measurements in those samples are not certified.

## 4. Apply an observed recovery rule

All four rule parameters are explicit. These illustrative choices are not
estimated from the baseline or recommended as biological thresholds:

| Parameter | Value | Meaning here |
|:----------|------:|:-------------|
| `threshold` | 0.25 | Within-band visits have deviation at most 0.25. |
| `persistence` | 4 | The supporting observed run must span at least four days. |
| `max_gap` | 3 | Consecutive supporting visits may be at most three days apart. |
| `horizon` | 10 | Detect and confirm within relative days 0 through 10. |

```{r observed-recovery}
rule <- list(threshold = 0.25, persistence = 4, max_gap = 3, horizon = 10)
recovered <- recoverome::add_recovery(tse, "observed", rule)
episode_results <- recoverome::recovery_results(recovered, "observed")
as.data.frame(episode_results)[, c(
  "status", "first_perturbation_time", "candidate_time", "confirmation_time",
  "rebound_time", "coverage"
)]
```

Day 0 has deviation `0.75`, above the threshold. Days 2, 4 and 6 are within
band and span four days, with two-day gaps. Day 2 is the candidate return;
day 6 supplies confirmation. Day 8 is a rebound, which preserves the first
confirmation. Coverage reaches the day-10 horizon.

```{r verify-outcome}
stopifnot(
  identical(episode_results$status, "confirmed_return"),
  identical(episode_results$first_perturbation_time, 0),
  identical(episode_results$candidate_time, 2),
  identical(episode_results$confirmation_time, 6),
  identical(episode_results$rebound_time, 8),
  identical(episode_results$coverage, "reaches_horizon")
)
context <- S4Vectors::metadata(episode_results)$recoverome_view
context$evidence$episode_1$confirmation_run
stopifnot(identical(
  context$evidence$episode_1$confirmation_run, c("s2", "s3", "s4")
))
```

Outcomes and evidence belong to episodes and are stored in metadata, with no
new recovery columns in `colData()`. The [observed rule contract][rfc3] defines
same-time grouping, run breaks and evaluation precedence. Same-time samples
must all be within band; an outside visit or excessive gap breaks a run.
Observations after the horizon contribute to coverage, not detection,
confirmation or rebound. Other registered events overlapping the analysis
interval can make the episode not evaluable.

Read status and coverage separately:

| Status | Interpretation |
|:-------|:---------------|
| `not_evaluable` | See `reason`: missing baseline, unresolved events or no observations in the window. |
| `no_detected_perturbation` | No above-threshold perturbation in the eligible detection interval. |
| `no_observed_return` | An observed perturbation has no later observed within-band return in the window. |
| `unconfirmed_return` | A return was observed without sufficient supporting persistence under the rule. |
| `confirmed_return` | A qualifying observed run supplies confirmation; inspect any later rebound. |

Coverage is `none`, `ends_before_horizon` or `reaches_horizon`. Confirmation
can occur before follow-up ends short of the horizon. Conversely, no observed
return before follow-up ends does not establish permanent non-recovery.

## 5. Validate dependencies and plot the complete analysis

Validation returns a plain list with `report_schema_version = 1L`, a `summary`
DataFrame and a `diagnostics` DataFrame. It checks stored records and relevant
current dependencies without modifying the TSE or recomputing the rule.

```{r validate-complete}
report <- recoverome::validate_recovery(recovered, "observed")
report$summary
unique(report$diagnostics$code)
stopifnot(
  report$summary$structural_valid,
  report$summary$validation_complete,
  report$summary$dependencies == "unchanged"
)
```

Limited single-baseline support is informational. Structural validity,
completion of checks and dependency changes are distinct fields, not a single
scientific-validity verdict. Registration-only validation does not read assays;
analytical stages compare relevant inputs, parent records and stored outputs.
The [diagnostic guide][validation] describes checks and detection boundaries.

```{r complete-plot, fig.width=9, fig.height=7, fig.alt="Seven observed deviations with return at day 2, confirmation at day 6 and rebound at day 8."}
p <- recoverome::plot_recovery(recovered, "observed")
p + ggplot2::labs(title = "Observed return, confirmation and rebound")
```

`plot_recovery()` returns an ordinary unprinted ggplot object. Customize it with
`+ ggplot2::labs(...)` or `+ ggplot2::theme(...)`. Facets identify the episode,
reference support, origin event and saved outcome. The separate evidence rail
shows milestones; its candidate-to-confirmation span represents observed
support, not continuously known recovery. Detection-window shading is a rule
parameter, not a confidence interval. Before `add_recovery()`, plotting a
completed deviation stage shows observations without an invented rule.

## 6. Filter without rewriting the completed analysis

Remove day-4 sample `s3`, an intermediate supporting visit. Ordinary TSE
subsetting preserves the historical reference and episode outcome. It does not
rerun the rule on the remaining visits or reconstruct removed deviations.

```{r filtered-analysis}
filtered <- recovered[, colnames(recovered) != "s3"]
filtered_report <- recoverome::validate_recovery(filtered, "observed")
filtered_report$summary

historical <- recoverome::recovery_results(filtered, "observed", scope = "historical")
historical[, c("result_state", "confirmation_time", "validation_complete", "dependencies")]
stopifnot(
  identical(historical$confirmation_time, 6),
  identical(historical$result_state, "available"),
  identical(historical$structural_valid, TRUE),
  identical(historical$validation_complete, FALSE),
  identical(historical$dependencies, "not_checked")
)
```

Confirmation remains available as a historical result, but missing inputs
prevent complete comparison with the original analysis. This is not a newly
evaluated unconfirmed return on the subset.

```{r historical-samples}
samples <- recoverome::recovery_results(
  filtered, "observed", level = "sample", scope = "historical"
)
as.data.frame(samples)[, c("sample_id", "relative_time", "result_state", "deviation")]
removed <- samples[samples$sample_id == "s3", ]
stopifnot(
  identical(removed$relative_time, 4),
  identical(removed$result_state, "removed"),
  identical(removed$deviation, NA_real_),
  identical(removed$deviation_status, NA_character_)
)
```

The [extraction and plotting contract][rfc4] keeps availability separate from
validation. `scope = "current"` selects retained original sample IDs or episodes
with retained included samples; `"historical"` selects the original scope.
Both use registration order and saved coordinates. Newly added IDs are not
enrolled. Validation columns describe the whole selected analysis, not each row.

Tables have atomic columns and convert using `as.data.frame()` or, optionally,
`tibble::as_tibble(as.data.frame(samples))`. The context in
`S4Vectors::metadata(historical)$recoverome_view` includes full validation,
registration, definitions, provenance and evidence. This is an extraction-time
snapshot: filtering or editing a returned table does not refresh it, and
conversion need not preserve its metadata. Extract again from the TSE for a
fresh view. Keep the TSE as the authoritative analysis.

```{r historical-plot, fig.width=9, fig.height=7, fig.alt="Historical confirmation remains at day 6, with hollow evidence markers and a dashed span after removal of the day-4 supporting visit."}
recoverome::plot_recovery(filtered, "observed", scope = "historical") +
  ggplot2::labs(title = "Historical confirmation after filtering")
```

The removed time appears on the evidence rail with no invented deviation.
Hollow milestones and dashed spans mean supporting observation IDs are missing,
not that timing is uncertain. Confirmation requires the entire saved run, so
its marker becomes hollow even though the day-6 endpoint sample remains.
Original evaluated gaps are preserved: filtering does not create a new original
gap or reclassify a run. Incomplete checks appear in the caption and facet.

For other sampling schedules, original gaps exceeding `max_gap` and follow-up
ending before the horizon have separate plot annotations. The unsampled region
does not establish an outside state. Historical episodes with no retained
values can still display their saved evidence.

### Changed inputs are different from unavailable inputs

An edit to a retained sample's day is a detectable source change. Extraction
keeps the original time and results, with changed dependency flags:

```{r changed-source}
changed <- recovered
annotations <- SummarizedExperiment::colData(changed)
annotations$day[colnames(changed) == "s4"] <- 30
SummarizedExperiment::colData(changed) <- annotations
changed_report <- recoverome::validate_recovery(changed, "observed")
changed_report$summary
saved <- recoverome::recovery_results(changed, "observed", level = "sample")
stopifnot(
  identical(saved$relative_time[saved$sample_id == "s4"], 6),
  all(saved$dependencies == "changed")
)
```

A changed comparison can be complete; an unavailable required comparison cannot.
Known changes take precedence over unavailable comparisons in the dependency
summary. Editing authoritative deviations or inconsistent parent records is a
structural problem: extraction errors rather than presenting corrupt outputs
as coherent results. Validation never repairs records. To ask a different
analytical question, create a new named analysis and explicitly calculate its
stages on suitable inputs.

## Optional tidy sample operations

The core examples above require no tidy adapter. The separate
[tidy interoperability assessment][tidy] verifies bounded filtering and
reordering directly on the existing TSE, using `tidySingleCellExperiment` 1.22.0
and dplyr 1.2.1 with TSE 2.20.0. No conversion to a different container is needed.
The `.cell` adapter column represents sample IDs, not a biological cell unit.

This optional recipe is not executed during the core vignette build because
those packages are not required dependencies. Its operations are covered by
the separate assessment; run it only in an environment with that adapter and
dplyr installed, without simultaneously loading other tidy adapters.

```{r optional-tidy, eval=FALSE}
stopifnot(requireNamespace("tidySingleCellExperiment", quietly = TRUE))
selected <- recovered |>
  dplyr::filter(.cell != "s3") |>
  dplyr::arrange(dplyr::desc(.cell))
recoverome::validate_recovery(selected, "observed")$summary
recoverome::recovery_results(selected, "observed", scope = "historical")
```

Mutation support is narrower: only ordinary atomic annotations with empty
`metadata(colData(tse))` and absent `mcols(colData(tse))` descriptors were
preserved. The tested adapter loses both forms of annotation information on
mutation, even for an unrelated new column. Use public `colData()` accessors
when those must survive. Sample identity mutation and row-taxonomy filtering
are unsupported in this view; other verbs and adapters remain unverified.
Filtering through a tidy method has the same historical interpretation as base
subsetting. It does not make the saved analysis current on the subset.

## Further examples, scope and session information

The bundled `single_episode` and `repeated_episodes` cases cover smaller
registration examples, multiple episodes for one subject, different event
boundaries and excluded samples. See `?recoverome::recovery_examples`, the
[reference examples][reference] and [history integration examples][history] for
these additional cases. All use explicit membership; reference selection never
borrows across episodes.

Statistical fitting, group comparisons and recovery-time uncertainty methods
remain future work requiring separate design and validation. The experimental
GitHub MVP target is `0.2.0`; this vignette does not publish that release or
claim Bioconductor acceptance. Record session versions alongside real analyses,
including optional adapters if used. Analysis records also retain their
creation-time package version and provenance.

```{r session-info}
utils::packageVersion("recoverome")
utils::sessionInfo()
```

[rfc1]: https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/001-registration-validation.md
[rfc2]: https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/002-personal-baseline-deviation.md
[rfc3]: https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/003-observed-recovery.md
[rfc4]: https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/004-result-extraction-plotting.md
[validation]: https://github.com/xec-cm/recoverome/blob/devel/dev/analytical-validation.md
[tidy]: https://github.com/xec-cm/recoverome/blob/devel/dev/tidy-interoperability.md
[reference]: https://github.com/xec-cm/recoverome/blob/devel/dev/examples/002-personal-baseline-deviation.R
[history]: https://github.com/xec-cm/recoverome/blob/devel/dev/recovery-history.md
