---
title: "Register an analysis and preserve its history"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Register an analysis and preserve its history}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Current status

The experimental development version provides `setup_recovery()`. This
vignette registers an analysis and inspects its stored history. Reference
estimation, deviation and recovery calculations, extraction, plotting, and
`validate_recovery()` remain planned.

## A small longitudinal input

The registration unit is an episode within a subject, with a perturbation
defined by events. Sample measurements and sample metadata remain in a
`TreeSummarizedExperiment` (TSE).

This deliberately small example has two features and three samples from one
subject. The time coordinate is days since enrolment, with an illustrative
exposure beginning on day 10. It demonstrates structure only; it cannot establish
sustained recovery or calibrate a statistical method.

```{r input-tse}
counts <- matrix(
  c(80L, 20L, 75L, 25L, 40L, 60L),
  nrow = 2L,
  dimnames = list(c("feature_a", "feature_b"), c("s1", "s2", "s3"))
)

sample_data <- S4Vectors::DataFrame(
  subject_id = rep("participant_1", 3L),
  episode_id = rep("episode_1", 3L),
  day = c(3, 10, 17),
  phase = c("before", "during", "after"),
  row.names = colnames(counts)
)

tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = counts),
  colData = sample_data
)

methods::validObject(tse)
SummarizedExperiment::assay(tse, "counts")
SummarizedExperiment::colData(tse)
```

A phylogenetic tree is not needed to construct this object. Any future method
that requires a tree will need to declare and validate that requirement.

## Declare episodes and events

An **episode** identifies the observation period relevant to one perturbation
and one subject. An **event** records a dated occurrence or interval, such as
the start and end of antibiotic exposure. Repeated perturbations need distinct
episode identities even when they occur in the same subject.

The episode selects the start or end of one of its events as its origin.
The tables below declare one episode and an exposure interval. Required
schemas and time rules are defined in the accepted
[RFC 001](https://github.com/xec-cm/recoverome/blob/devel/dev/rfcs/001-registration-validation.md).

```{r registration-inputs}
episodes <- data.frame(
  episode_id = "episode_1",
  subject_id = "participant_1",
  origin_event_id = "exposure_1",
  origin_boundary = "start"
)
events <- data.frame(
  event_id = "exposure_1",
  episode_id = "episode_1",
  start_time = 10,
  end_time = 14
)
episodes
events
```

Each sample is explicitly assigned to an episode by its `episode_id` column.
An `NA` episode explicitly excludes that sample from this named analysis;
each declared episode must still have an included sample. Registration accepts
finite numeric times with a declared unit and coordinate origin. Calendar
times need an explicit conversion by the caller before registration.

## Register and inspect the snapshot

`setup_recovery()` checks the identities, episode membership, and event
references, then returns the TSE with a named registration. Analysis IDs match
`^[a-z][a-z0-9]*$`. The corresponding `rec_<analysis>_` prefix is reserved for
future result columns; registration creates none.

```{r register-analysis}
tse <- recoverome::setup_recovery(
  tse,
  analysis_id = "antibiotic",
  episodes = episodes,
  events = events,
  time_col = "day",
  time_unit = "days",
  time_origin = "days since enrolment within each participant"
)

registration <- S4Vectors::metadata(tse)$recoverome$analyses$antibiotic
registration$registration$samples
registration$episodes
registration$events
registration$scope
registration$owned_columns
```

The stored tables are `S4Vectors::DataFrame` objects. The sample snapshot
contains included samples in their original order; the scope records all
original sample and feature IDs, including any excluded samples. This is
historical metadata, and must not be edited as a second current data source.
Registration also records its source column names, the time declaration, the
package version, and a UTC registration timestamp.

Here the event origin is day 10, so the implied relative sample times are
`-7`, `0`, and `7` days. Registration retains the supplied coordinates without
creating a second time column. It neither selects an assay nor reads or
fingerprints its values.

Calling setup again with the same analysis ID is an error, even with identical
inputs. A pre-existing column beginning with `rec_antibiotic_` also blocks this
registration. Another unused name can register an independent analysis.

## Filtering preserves analysis history

Ordinary TSE subsetting selects features, samples, and their annotations while
preserving the original registration. This subset contains one feature and
only the sample collected after the exposure:

```{r subset-tse}
after_event <- tse["feature_a", "s3", drop = FALSE]
rownames(after_event)
colnames(after_event)
history <- S4Vectors::metadata(after_event)$recoverome$analyses$antibiotic
history$scope
identical(history, registration)
```

The stored scope still identifies both features and all three samples, and the
entire registration is unchanged. The snapshot describes the original data;
filtering does not make it a new registration of the subset.

The planned `validate_recovery()` will distinguish reduced scope from changes
to consumed sample metadata. It is not available in this version. Registration
alone makes no claim about the suitability of assay values for a later method.

## Planned analytical stages

A **reference** will record the comparison target and eligible samples; a
recovery rule will define an acceptable deviation and observational requirements.
Their analytical definitions require separate design decisions. The following
is **non-executable proposed API**; these six functions are not implemented.

```{r planned-workflow, eval=FALSE}
tse <- add_reference(tse, analysis_id = "antibiotic", reference = reference)
tse <- add_deviation(tse, analysis_id = "antibiotic")
tse <- add_recovery(tse, analysis_id = "antibiotic", rule = recovery_rule)

validate_recovery(tse, analysis_id = "antibiotic")
recovery_results(tse, analysis_id = "antibiotic", level = "episode")
plot_recovery(tse, analysis_id = "antibiotic")
```

The add functions are intended to return a TSE. Extraction, plotting, and
validation have separate roles and must not silently alter an analysis.
Statistical fitting is a future layer outside this initial interface.

## Interpretation boundaries

Observed similarity at one visit does not establish a sustained return. A rule
based on repeated visits must also state the relevant duration and admissible
gaps. Sparse observations cannot establish continuous behavior between visits
without additional assumptions.

Likewise, the absence of an observed recovery by the end of follow-up does not
establish that recovery will never occur. Methods for uncertainty, incomplete
follow-up, and comparisons between groups will need dedicated statistical
validation before implementation claims can be made.

The intended outcome is an explicit, auditable description of recovery under
a stated definition. Recovery to baseline is not a clinical diagnosis or a
claim that all microbial functions have returned.

## Session information

```{r session-info}
sessionInfo()
```
