---
title: "Compare explicit recovery rules"
author: "Francesc Catala-Moll"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Compare explicit recovery rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(recoverome)
```

## Introduction

A recovery outcome depends on an explicit band, persistence duration, maximum
visit gap and horizon. `recovery_sensitivity()` compares settings you supply,
using a fixed personal reference and saved deviations. It shows the consequences
of those choices without selecting a best setting or replacing your original
analysis. Install as described in [Get started](recoverome.html).

Choose plausible alternatives before examining which settings produce a return.
Baseline and sampling sensitivity require new named analyses; this API varies
only the four observation-rule parameters. See the [real-data assessment](real-data.html)
for why arithmetic correctness and observation adequacy are separate questions.

## Prepare one analysis and its original outcome

```{r original-analysis}
data("recovery_examples", package = "recoverome")
example_data <- recovery_examples$observed_recovery
tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = example_data$counts),
  colData = S4Vectors::DataFrame(example_data$col_data)
)
tse <- setup_recovery(
  tse, "observed", example_data$episodes, example_data$events,
  time_col = "day", time_unit = "days", time_origin = "days since enrolment"
)
tse <- add_reference(tse, "observed", "b1", assay = "counts")
tse <- add_deviation(tse, "observed")
tse <- add_recovery(tse, "observed", list(
  threshold = .25, persistence = 4, max_gap = 3, horizon = 10
))
before <- serialize(tse, NULL)
```

The bundled synthetic example has known deviations of .75, .25, .125, .125, .5
and .125 at follow-up days 0, 2, 4, 6, 8 and 10. Its original outcome confirms at
6 and records a rebound at 8. The following illustrative alternatives are not
recommended biological parameters.

## Supply every scenario explicitly

```{r scenarios}
rules <- data.frame(
  scenario_id = c("primary", "short_gap", "tight_band", "short_horizon"),
  threshold = c(.25, .25, .10, .25),
  persistence = 4,
  max_gap = c(3, 1, 3, 3),
  horizon = c(10, 10, 10, 4)
)
sensitivity <- recovery_sensitivity(tse, "observed", rules)
as.data.frame(sensitivity)[, c(
  "scenario_id", "evaluation_state", "status", "candidate_time", "confirmation_time", "rebound_time"
)]
stopifnot(
  identical(serialize(tse, NULL), before),
  identical(sensitivity$status, c("confirmed_return", "unconfirmed_return",
                                  "no_observed_return", "unconfirmed_return")),
  identical(sensitivity$confirmation_time, c(6, NA_real_, NA_real_, NA_real_))
)
```

Shortening the allowed gap to one day prevents confirmation because the visits
are two days apart. Tightening the band to .10 gives no observed return: the
smallest follow-up deviation is .125. Ending the horizon at day 4 observes a
return but not enough persistence. None of these results should be hidden in
favor of the confirmed setting. Confirmation still does not establish continuous
or permanent recovery between measurements.

```{r scenario-plot, fig.width=9, fig.height=6, fig.alt="Four explicitly supplied rules produce one confirmed return, two unconfirmed returns and one no-observed-return outcome."}
plot_sensitivity(sensitivity)
```

Rows preserve supplied scenario order; columns identify both participant and
episode. Cell labels give observed first return, confirmation and rebound, while
colors indicate the categorical outcome. This is a comparison of definitions,
not uncertainty intervals or a test for an optimal rule. The result is an editable
ggplot; enlarge the device for many scenarios or episodes.

## Keep unavailable scenarios visible

```{r historical-inputs}
filtered <- tse[, colnames(tse) != "s3"]
unavailable <- recovery_sensitivity(filtered, "observed", rules)
as.data.frame(unavailable)[, c(
  "scenario_id", "evaluation_state", "evaluation_reason", "status", "confirmation_time"
)]
stopifnot(all(unavailable$evaluation_reason == "historical_inputs_unavailable"),
          all(is.na(unavailable$confirmation_time)))
recovery_results(filtered, "observed")$confirmation_time
```

The original historical confirmation remains 6, but it is **not copied into the
new scenarios**. The removed intermediate visit is required to re-evaluate the
rule. Changed consumed inputs give `inputs_changed`; otherwise unsupported or
incomplete comparisons give `validation_incomplete`. These reasons are distinct
from a scenario that was evaluated and returned `not_evaluable`, for example
because its episode has no baseline or overlapping exposures.

```{r unavailable-plot, fig.width=9, fig.height=6, fig.alt="All four requested scenarios remain visible as unavailable because an original supporting sample was removed."}
plot_sensitivity(unavailable)
```

All originally registered episodes remain represented, even if one has no samples
left in the current TSE. `current_present` distinguishes that case. Removing only
features not used by the reference, or reordering samples, does not itself prevent
evaluation. Added identities are not enrolled. Malformed rules or corrupted
required records error instead of producing a success-shaped partial result.

## Preserve context when working with tables

```{r context}
context <- S4Vectors::metadata(sensitivity)$recoverome_sensitivity
context$rules
context$validation$summary
context$evidence$primary$episode_1$confirmation_run
```

The DataFrame contains atomic columns suitable for conversion to an ordinary
data.frame or tibble. Its metadata records rules, validation, the original analysis
context, scenario-specific supporting IDs and provenance. This is a snapshot, not
a live link to the TSE. Use a separate converted copy for table manipulation.
`plot_sensitivity()` expects the unchanged result and verifies its versioned
consistency checksum; edits or row filtering must not silently mismatch the saved
scenario context. A fresh analysis call makes a fresh snapshot.

## Session information

```{r session-info}
utils::sessionInfo()
```
